---
title: "Indonesia Inequality Project: EDA"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

This Quarto notebook mirrors the Python EDA with a story-first structure. The main section keeps only the charts needed for the current narrative. Validation checks and supporting charts are kept in the appendix.

```{r setup}
library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

project_root <- if (basename(getwd()) == "notebooks") dirname(getwd()) else getwd()
data_dir <- file.path(project_root, "data", "processed")

palette <- c(
  blue = "#2F6B9A",
  red = "#C44E52",
  green = "#4C956C",
  gold = "#D4A017",
  purple = "#7D5BA6",
  gray = "#6C757D",
  light = "#F6F7F9",
  dark = "#1F2933"
)

chart_theme <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14, color = palette[["dark"]]),
      plot.subtitle = element_text(size = 10, color = palette[["gray"]]),
      axis.title = element_text(size = 10, color = palette[["dark"]]),
      axis.text = element_text(size = 9, color = palette[["dark"]]),
      legend.title = element_text(size = 9, face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

read_processed_csv <- function(filename) {
  read_csv(file.path(data_dir, filename), show_col_types = FALSE) %>%
    rename_with(str_trim)
}
```

## Load Data

```{r load-data}
files <- c(
  gdp = "gdp_indonesia.csv",
  inflation = "inflation_indonesia.csv",
  inequality = "inequality_indonesia.csv",
  welfare_distribution = "welfare_distribution_indonesia.csv",
  comparative_growth = "comparative_inequality_growth.csv",
  poverty = "poverty_indonesia.csv",
  labor = "labor_indonesia.csv",
  cost_of_living = "cost_of_living_indonesia.csv"
)

data <- files %>%
  imap(~ read_processed_csv(.x)) %>%
  set_names(names(files))

indonesia_data <- data[names(data) != "comparative_growth"]
```

# Main Story

The storyline is: Indonesia is perceived as increasingly unequal, real GDP per capita has risen, average consumption improved across broad groups, but the distribution of gains shifted toward higher-income groups. The final section asks whether households feel better off once wages are compared with prices.

## Indonesia's Relative Inequality Position

```{r comparative-gini-position, fig.width=10, fig.height=6}
comparison_countries <- tibble(
  country_code = c(
    "IDN", "VNM", "THA", "MYS", "PHL", "CHN", "IND",
    "BRA", "MEX", "CHL",
    "JPN", "KOR", "DEU", "FRA", "AUS", "USA"
  ),
  country_label = c(
    "Indonesia", "Vietnam", "Thailand", "Malaysia", "Philippines", "China", "India",
    "Brazil", "Mexico", "Chile",
    "Japan", "South Korea", "Germany", "France", "Australia", "United States"
  ),
  comparison_group = c(
    rep("Asian peers", 7),
    rep("High-inequality emerging references", 3),
    rep("Advanced-country benchmarks", 6)
  )
)

latest_gini_comparison <- data$comparative_growth %>%
  inner_join(comparison_countries, by = "country_code") %>%
  drop_na(gini) %>%
  group_by(country_code) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    country_label = reorder(country_label, gini),
    is_indonesia = country_code == "IDN",
    label = paste0(round(gini, 1), " (", year, ")")
  )

latest_gini_comparison %>%
  ggplot(aes(country_label, gini, fill = is_indonesia)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = label), hjust = -0.08, size = 3) +
  coord_flip() +
  facet_wrap(~ comparison_group, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c(`FALSE` = palette[["gray"]], `TRUE` = palette[["red"]])) +
  scale_y_continuous(limits = c(0, 62), breaks = seq(0, 60, by = 10)) +
  labs(
    title = "Latest Available Gini Index: Indonesia and Comparator Countries",
    subtitle = "World Bank; labels show Gini value and observation year. Higher Gini means higher measured inequality.",
    x = NULL, y = "Gini index"
  ) +
  chart_theme() +
  theme(legend.position = "none")
```

## Real GDP Per Capita

```{r gdp-per-capita, fig.width=10, fig.height=5.5}
data$gdp %>%
  drop_na(real_gdp_per_capita_lcu) %>%
  ggplot(aes(year, real_gdp_per_capita_lcu)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Real GDP per Capita",
    subtitle = "World Bank API, constant local currency; price effects already removed",
    x = "Year", y = "Constant Indonesian rupiah"
  ) +
  chart_theme()
```

## Average Consumption By Population Group

```{r group-consumption, fig.width=10, fig.height=6}
broad_groups <- c("Bottom 40%", "Middle 40%", "Top 20%")

group_consumption <- data$welfare_distribution %>%
  filter(group %in% broad_groups) %>%
  mutate(group = factor(group, levels = broad_groups)) %>%
  drop_na(avg_welfare_ppp_daily)

group_consumption_labels <- group_consumption %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_welfare = first(avg_welfare_ppp_daily)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (avg_welfare_ppp_daily / first_welfare - 1) * 100,
    growth_label = sprintf("+%.0f%%", growth_pct)
  )

group_consumption %>%
  ggplot(aes(year, avg_welfare_ppp_daily, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = group_consumption_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c(palette[["green"]], palette[["blue"]], palette[["red"]])) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Average Consumption by Population Group",
    subtitle = "End labels show growth from first available year to final year; 2017 PPP USD per person per day",
    x = "Year", y = "Daily consumption", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

## Distribution Shares By Population Group

```{r distribution-shares, fig.width=10, fig.height=5.5}
ineq_long <- data$welfare_distribution %>%
  filter(group %in% broad_groups) %>%
  mutate(
    group = factor(group, levels = broad_groups),
    share = welfare_share * 100
  ) %>%
  drop_na(share)

ineq_change_labels <- ineq_long %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    change_pp = share - first_share,
    change_label = sprintf("%+.1f pp", change_pp)
  )

ineq_long %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = ineq_change_labels,
    aes(label = change_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(palette[["green"]], palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Distribution Shares by Population Group",
    subtitle = "End labels show percentage-point change from first available year to final year",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 70, 5.5, 5.5))
```

## Tail Distribution Shares

```{r tail-distribution-shares, fig.width=10, fig.height=5.5}
tail <- data$welfare_distribution %>%
  filter(group %in% c("Bottom 10%", "Top 10%")) %>%
  mutate(group = factor(group, levels = c("Bottom 10%", "Top 10%")))

tail_shares <- tail %>%
  select(year, group, welfare_share) %>%
  mutate(share = welfare_share * 100) %>%
  drop_na(share)

tail_share_labels <- tail_shares %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (share / first_share - 1) * 100,
    growth_label = sprintf("%+.0f%%", growth_pct)
  )

tail_shares %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = tail_share_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Bottom 10%" = palette[["green"]],
    "Top 10%" = palette[["red"]]
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Tail Distribution Shares",
    subtitle = "End labels show growth from first available year to final year; World Bank PIP consumption shares",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

## Inflation And CPI

```{r inflation-cpi, fig.width=10, fig.height=5.5}
inflation_long <- data$inflation %>%
  select(year, cpi_total, inflation_yoy) %>%
  pivot_longer(
    cols = c(cpi_total, inflation_yoy),
    names_to = "metric",
    values_to = "value"
  ) %>%
  drop_na(value) %>%
  mutate(metric = recode(
    metric,
    cpi_total = "CPI total index",
    inflation_yoy = "Inflation YoY (%)"
  ))

inflation_long %>%
  ggplot(aes(year, value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c(palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Inflation and CPI Coverage Available Now",
    subtitle = "Total CPI and annual inflation from World Bank",
    x = "Year", y = NULL, color = "Metric"
  ) +
  chart_theme()
```

## BPS Category CPI Indexes

```{r bps-category-cpi, fig.width=10, fig.height=6}
category_cpi <- data$inflation %>%
  select(year, cpi_food, cpi_transport, cpi_housing, cpi_education, cpi_healthcare, cpi_restaurant) %>%
  pivot_longer(
    cols = starts_with("cpi_"),
    names_to = "category",
    values_to = "index"
  ) %>%
  drop_na(index) %>%
  mutate(category = recode(
    category,
    cpi_food = "Food, beverages, tobacco",
    cpi_transport = "Transport",
    cpi_housing = "Housing, water, electricity, fuel",
    cpi_education = "Education",
    cpi_healthcare = "Health",
    cpi_restaurant = "Restaurants"
  ))

category_cpi %>%
  ggplot(aes(year, index, color = category)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~ category, scales = "free_y", ncol = 3) +
  scale_x_continuous(breaks = 2020:2023) +
  scale_color_manual(values = rep(c(
    palette[["green"]],
    palette[["blue"]],
    palette[["red"]],
    palette[["gold"]],
    palette[["purple"]],
    palette[["gray"]]
  ), length.out = length(unique(category_cpi$category)))) +
  labs(
    title = "BPS Category CPI Indexes",
    subtitle = "National December indexes, 2018=100 base; 2024 uses a new BPS base and is not stitched here",
    x = "Year", y = "Index", color = "Category"
  ) +
  chart_theme()
```

## Real Wage Growth

```{r real-wage-growth, fig.width=10, fig.height=5.5}
wage_growth <- data$labor %>%
  select(year, avg_wage, real_wage_growth) %>%
  drop_na(avg_wage)

wage_growth %>%
  ggplot(aes(year, real_wage_growth)) +
  geom_hline(yintercept = 0, color = palette[["gray"]], linetype = "dashed") +
  geom_col(aes(fill = real_wage_growth >= 0), width = 0.75) +
  scale_fill_manual(values = c(`FALSE` = palette[["red"]], `TRUE` = palette[["green"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Real Wage Growth",
    subtitle = "Average monthly employee earnings from ILOSTAT, deflated with World Bank total CPI",
    x = "Year", y = "Real wage growth (%)"
  ) +
  chart_theme() +
  theme(legend.position = "none")
```

## Nominal Wages, Prices, And Purchasing Power

```{r wage-vs-cpi-index, fig.width=10, fig.height=5.5}
wage_price_index <- data$labor %>%
  select(year, avg_wage) %>%
  left_join(data$inflation %>% select(year, cpi_total), by = "year") %>%
  filter(year >= 2019) %>%
  drop_na(avg_wage, cpi_total) %>%
  mutate(real_wage = avg_wage / cpi_total * 100) %>%
  mutate(
    nominal_wage_index = avg_wage / first(avg_wage) * 100,
    cpi_index = cpi_total / first(cpi_total) * 100,
    real_wage_index = real_wage / first(real_wage) * 100
  ) %>%
  select(year, nominal_wage_index, cpi_index, real_wage_index) %>%
  pivot_longer(
    cols = ends_with("_index"),
    names_to = "metric",
    values_to = "index"
  ) %>%
  mutate(metric = recode(
    metric,
    nominal_wage_index = "Nominal wage",
    cpi_index = "Consumer prices",
    real_wage_index = "Real wage"
  ))

wage_price_index %>%
  ggplot(aes(year, index, color = metric)) +
  geom_hline(yintercept = 100, color = palette[["gray"]], linetype = "dashed") +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = c(
    "Nominal wage" = palette[["blue"]],
    "Consumer prices" = palette[["red"]],
    "Real wage" = palette[["green"]]
  )) +
  scale_x_continuous(breaks = 2019:2024) +
  labs(
    title = "Nominal Wages, Prices, and Purchasing Power",
    subtitle = "Indexed to 2019 = 100; real wage is nominal wage divided by CPI",
    x = "Year", y = "Index, 2019 = 100", color = "Metric"
  ) +
  chart_theme()
```

## Household Price Pressure

```{r household-price-pressure, fig.width=10, fig.height=6}
price_pressure <- bind_rows(
  data$inflation %>%
    select(year, cpi_total) %>%
    mutate(metric = "Total CPI", value = cpi_total) %>%
    select(year, metric, value),
  category_cpi %>%
    transmute(year, metric = category, value = index),
  data$cost_of_living %>%
    select(year, rice_price) %>%
    mutate(metric = "Wholesale rice price", value = rice_price) %>%
    select(year, metric, value),
  data$labor %>%
    select(year, avg_wage) %>%
    mutate(metric = "Average monthly earnings", value = avg_wage) %>%
    select(year, metric, value)
) %>%
  filter(year >= 2020) %>%
  drop_na(value) %>%
  group_by(metric) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    base_value = first(value),
    index_2020 = value / base_value * 100
  ) %>%
  ungroup()

price_pressure %>%
  ggplot(aes(year, index_2020, color = metric)) +
  geom_hline(yintercept = 100, color = palette[["gray"]], linetype = "dashed") +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y", ncol = 3) +
  scale_x_continuous(breaks = 2020:2024) +
  labs(
    title = "How Prices Move From a Household Starting Point",
    subtitle = "Each series rebased to 2020 = 100. If earnings rise more slowly than prices, purchasing power falls.",
    x = "Year", y = "Index, 2020 = 100", color = "Series"
  ) +
  chart_theme() +
  theme(legend.position = "none")
```

# Appendix

## Data Validation

```{r data-summary}
data %>%
  imap_dfr(~ tibble(
    dataset = .y,
    rows = nrow(.x),
    first_year = min(.x$year, na.rm = TRUE),
    last_year = max(.x$year, na.rm = TRUE),
    columns = ncol(.x)
  ))
```

```{r data-coverage, fig.width=10, fig.height=5.5}
coverage <- indonesia_data %>%
  imap_dfr(function(df, dataset_name) {
    value_cols <- df %>%
      select(-any_of(c("year", "source"))) %>%
      names()

    df %>%
      rowwise() %>%
      mutate(
        dataset = dataset_name,
        available_fields = sum(!is.na(pick(all_of(value_cols)))),
        total_fields = length(value_cols),
        coverage_rate = if_else(total_fields > 0, available_fields / total_fields, 0)
      ) %>%
      ungroup() %>%
      select(dataset, year, available_fields, total_fields, coverage_rate)
  })

coverage %>%
  ggplot(aes(year, dataset, fill = coverage_rate)) +
  geom_tile(color = "white", linewidth = 0.25) +
  scale_fill_gradient(low = "#F1F3F5", high = palette[["blue"]], labels = percent_format()) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Processed Data Coverage by Dataset and Year",
    subtitle = "Darker cells have more populated fields in the standardized CSV schema",
    x = "Year", y = NULL, fill = "Coverage"
  ) +
  chart_theme()
```

```{r field-availability-table}
availability <- indonesia_data %>%
  imap_dfr(function(df, dataset_name) {
    df %>%
      select(-any_of(c("year", "source"))) %>%
      summarise(across(everything(), ~ sum(!is.na(.x)))) %>%
      pivot_longer(everything(), names_to = "field", values_to = "non_missing_count") %>%
      mutate(
        dataset = dataset_name,
        total_rows = nrow(df),
        status = case_when(
          non_missing_count == 0 ~ "Empty",
          non_missing_count == total_rows ~ "Complete",
          TRUE ~ "Partly missing"
        )
      )
  }) %>%
  mutate(status = factor(status, levels = c("Complete", "Partly missing", "Empty")))

availability %>%
  arrange(dataset, status, field)
```

```{r field-availability-chart, fig.width=10, fig.height=5.5}
availability %>%
  count(dataset, status, name = "field_count") %>%
  ggplot(aes(dataset, field_count, fill = status)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c(
    Complete = palette[["green"]],
    `Partly missing` = palette[["gold"]],
    Empty = palette[["red"]]
  )) +
  labs(
    title = "Field Availability by Dataset",
    subtitle = "Empty fields are mostly waiting for source mapping or API recovery",
    x = NULL, y = "Number of fields", fill = "Status"
  ) +
  chart_theme() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r empty-fields}
availability %>%
  filter(status == "Empty") %>%
  select(dataset, field) %>%
  arrange(dataset, field)
```

```{r combined-dataset}
combined <- data$gdp %>%
  select(-source)

for (name in c("inflation", "inequality", "poverty", "labor", "cost_of_living")) {
  df <- data[[name]] %>%
    select(-source)

  duplicate_cols <- intersect(names(df), names(combined)) %>%
    setdiff("year")

  df <- df %>%
    select(-any_of(duplicate_cols))

  combined <- combined %>%
    full_join(df, by = "year")
}

combined <- combined %>%
  arrange(year)

combined %>%
  head()
```

## Supporting Charts

```{r gdp-growth, fig.width=10, fig.height=5.5}
data$gdp %>%
  drop_na(gdp_growth) %>%
  ggplot(aes(year, gdp_growth)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = palette[["gray"]]) +
  geom_line(color = palette[["blue"]], linewidth = 1.1) +
  geom_point(color = palette[["blue"]], size = 1.8) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "GDP Growth, Year over Year",
    subtitle = "World Bank API, annual percent",
    x = "Year", y = "GDP growth (%)"
  ) +
  chart_theme()
```

```{r tail-absolute-consumption, fig.width=10, fig.height=5.5}
tail_consumption <- tail %>%
  select(year, group, avg_welfare_ppp_daily) %>%
  drop_na(avg_welfare_ppp_daily)

tail_consumption_labels <- tail_consumption %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_welfare = first(avg_welfare_ppp_daily)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (avg_welfare_ppp_daily / first_welfare - 1) * 100,
    growth_label = sprintf("+%.0f%%", growth_pct)
  )

tail_consumption %>%
  ggplot(aes(year, avg_welfare_ppp_daily, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = tail_consumption_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Bottom 10%" = palette[["green"]],
    "Top 10%" = palette[["red"]]
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Tail Average Consumption",
    subtitle = "End labels show growth from first available year to final year; 2017 PPP USD per person per day",
    x = "Year", y = "Daily consumption", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

```{r extreme-tail-distribution-shares, fig.width=10, fig.height=5.5}
extreme_tail <- data$welfare_distribution %>%
  filter(group %in% c("Bottom 1%", "Top 1%")) %>%
  mutate(
    group = factor(group, levels = c("Bottom 1%", "Top 1%")),
    share = welfare_share * 100
  ) %>%
  drop_na(share)

extreme_tail_share_labels <- extreme_tail %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (share / first_share - 1) * 100,
    growth_label = sprintf("%+.0f%%", growth_pct)
  )

extreme_tail %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = extreme_tail_share_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Bottom 1%" = palette[["green"]],
    "Top 1%" = palette[["red"]]
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Extreme Tail Distribution Shares",
    subtitle = "End labels show growth from first available year to final year; World Bank PIP consumption shares",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

```{r gini-index, fig.width=10, fig.height=5.5}
data$inequality %>%
  drop_na(gini_ratio) %>%
  ggplot(aes(year, gini_ratio)) +
  geom_line(color = palette[["purple"]], linewidth = 1.1) +
  geom_point(color = palette[["purple"]], size = 1.8) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Gini Index",
    subtitle = "World Bank API, Indonesia",
    x = "Year", y = "Gini index"
  ) +
  chart_theme()
```

```{r comparative-inequality-growth, fig.width=10, fig.height=5.8}
comparative_long <- data$comparative_growth %>%
  filter(income_level %in% c("Lower middle income", "Upper middle income")) %>%
  select(
    country_code, country_name, region, income_level, year, gini,
    future_5y_growth_pct, future_10y_growth_pct
  ) %>%
  pivot_longer(
    cols = c(future_5y_growth_pct, future_10y_growth_pct),
    names_to = "horizon",
    values_to = "future_growth_pct"
  ) %>%
  mutate(
    horizon = recode(
      horizon,
      future_5y_growth_pct = "Next 5 years",
      future_10y_growth_pct = "Next 10 years"
    ),
    focus = if_else(country_code == "IDN", "Indonesia", "Other middle-income countries")
  ) %>%
  drop_na(gini, future_growth_pct)

comparative_long %>%
  ggplot(aes(gini, future_growth_pct)) +
  geom_hline(yintercept = 0, color = palette[["gray"]], linetype = "dashed") +
  geom_point(
    data = comparative_long %>% filter(focus != "Indonesia"),
    color = palette[["gray"]],
    alpha = 0.28,
    size = 1.7
  ) +
  geom_smooth(
    data = comparative_long %>% filter(focus != "Indonesia"),
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = palette[["purple"]],
    linewidth = 1
  ) +
  geom_point(
    data = comparative_long %>% filter(focus == "Indonesia"),
    color = palette[["red"]],
    size = 2.4
  ) +
  facet_wrap(~ horizon, ncol = 2) +
  scale_x_continuous(limits = c(20, 65), breaks = seq(20, 65, by = 10)) +
  labs(
    title = "Inequality Today and Future Growth",
    subtitle = "Middle-income country-years; red points highlight Indonesia. Trend line is descriptive, not causal.",
    x = "Gini index in year t",
    y = "Annualized future real GDP per-capita growth (%)"
  ) +
  chart_theme()
```

```{r poverty-indicators, fig.width=10, fig.height=5.5}
poverty_long <- data$poverty %>%
  select(year, poverty_rate, extreme_poverty_rate) %>%
  pivot_longer(
    cols = c(poverty_rate, extreme_poverty_rate),
    names_to = "metric",
    values_to = "value"
  ) %>%
  drop_na(value) %>%
  mutate(metric = recode(
    metric,
    poverty_rate = "National poverty rate",
    extreme_poverty_rate = "Extreme poverty rate"
  ))

poverty_long %>%
  ggplot(aes(year, value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = c(palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Poverty Indicators",
    subtitle = "World Bank API, annual percent of population",
    x = "Year", y = "Population share (%)", color = "Metric"
  ) +
  chart_theme()
```

```{r labor-market, fig.width=10, fig.height=5.5}
labor_long <- data$labor %>%
  select(year, unemployment_rate, youth_unemployment, vulnerable_employment_share) %>%
  pivot_longer(
    cols = c(unemployment_rate, youth_unemployment, vulnerable_employment_share),
    names_to = "metric",
    values_to = "value"
  ) %>%
  drop_na(value) %>%
  mutate(metric = recode(
    metric,
    unemployment_rate = "Unemployment",
    youth_unemployment = "Youth unemployment",
    vulnerable_employment_share = "Vulnerable employment"
  ))

labor_long %>%
  ggplot(aes(year, value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c(palette[["blue"]], palette[["red"]], palette[["green"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Labor Market Indicators Available Now",
    subtitle = "Unemployment and vulnerable employment from World Bank",
    x = "Year", y = "Share (%)", color = "Metric"
  ) +
  chart_theme()
```

```{r rice-price, fig.width=10, fig.height=5.5}
data$cost_of_living %>%
  drop_na(rice_price) %>%
  ggplot(aes(year, rice_price)) +
  geom_line(color = palette[["green"]], linewidth = 1.1) +
  geom_point(color = palette[["green"]], size = 1.8) +
  scale_x_continuous(breaks = seq(2010, 2024, by = 2)) +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(
    title = "Wholesale Rice Price From BPS",
    subtitle = "BPS WebAPI variable 295; rupiah per kilogram, annual average",
    x = "Year", y = "Rupiah per kg"
  ) +
  chart_theme()
```

```{r wage-price-summary}
data$labor %>%
  select(year, avg_wage, real_wage_growth) %>%
  left_join(data$inflation %>% select(year, inflation_yoy), by = "year") %>%
  filter(year >= 2019, year <= 2023) %>%
  mutate(
    nominal_wage_growth = (avg_wage / lag(avg_wage) - 1) * 100,
    real_wage_growth = round(real_wage_growth, 2),
    nominal_wage_growth = round(nominal_wage_growth, 2),
    inflation_yoy = round(inflation_yoy, 2)
  ) %>%
  select(year, avg_wage, nominal_wage_growth, inflation_yoy, real_wage_growth)
```

```{r cost-of-living-missingness, fig.width=10, fig.height=5.5}
data$cost_of_living %>%
  select(-any_of(c("year", "source"))) %>%
  summarise(across(everything(), ~ mean(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "field", values_to = "missing_rate") %>%
  ggplot(aes(missing_rate, field)) +
  geom_col(fill = palette[["gold"]]) +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Cost-of-Living Field Missingness",
    subtitle = "Total CPI and BPS wholesale rice price are populated; remaining fields need source mapping",
    x = "Share of rows missing", y = NULL
  ) +
  chart_theme()
```

## Next EDA Tasks

- Validate ILOSTAT average monthly earnings against BPS/Sakernas wage tables if a cleaner national API series is confirmed.
- Extend CPI category history before 2020 and bridge the 2024 BPS 2022=100 base only if the bridge is source-backed.
- Extend cost-of-living charts beyond wholesale rice once fuel/electricity/LPG/rent source IDs are mapped.
- Replace provisional distribution shares with BPS/SUSENAS expenditure-only distribution if available.
- Add informal employment charts once BPS or ILOSTAT source series are confirmed.
