---
title: "Indonesia Inequality Project: EDA"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

This Quarto notebook mirrors the Python EDA with a story-first structure. The main section keeps only the charts needed for the current narrative. Validation checks and supporting charts are kept in the appendix.

```{r setup}
library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

project_root <- if (basename(getwd()) == "notebooks") dirname(getwd()) else getwd()
data_dir <- file.path(project_root, "data", "processed")

palette <- c(
  blue = "#2F6B9A",
  red = "#C44E52",
  green = "#4C956C",
  gold = "#D4A017",
  purple = "#7D5BA6",
  gray = "#6C757D",
  light = "#F6F7F9",
  dark = "#1F2933"
)

chart_theme <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14, color = palette[["dark"]]),
      plot.subtitle = element_text(size = 10, color = palette[["gray"]]),
      axis.title = element_text(size = 10, color = palette[["dark"]]),
      axis.text = element_text(size = 9, color = palette[["dark"]]),
      legend.title = element_text(size = 9, face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

read_processed_csv <- function(filename) {
  read_csv(file.path(data_dir, filename), show_col_types = FALSE) %>%
    rename_with(str_trim)
}
```

## Load Data

```{r load-data}
files <- c(
  gdp = "gdp_indonesia.csv",
  inflation = "inflation_indonesia.csv",
  inequality = "inequality_indonesia.csv",
  welfare_distribution = "welfare_distribution_indonesia.csv",
  comparative_growth = "comparative_inequality_growth.csv",
  poverty = "poverty_indonesia.csv",
  labor = "labor_indonesia.csv",
  cost_of_living = "cost_of_living_indonesia.csv"
)

data <- files %>%
  imap(~ read_processed_csv(.x)) %>%
  set_names(names(files))

indonesia_data <- data[names(data) != "comparative_growth"]
```

# Main Story

The storyline is: Indonesia is perceived as increasingly unequal, real GDP per capita has risen, average consumption improved across broad groups, but the distribution of gains shifted toward higher-income groups. The final section asks whether households feel better off once wages are compared with prices.

## Indonesia's Relative Inequality Position

```{r comparative-gini-position, fig.width=10, fig.height=6}
comparison_countries <- tibble(
  country_code = c(
    "IDN", "VNM", "THA", "MYS", "PHL", "CHN", "IND",
    "BRA", "MEX", "CHL",
    "JPN", "KOR", "DEU", "FRA", "AUS", "USA"
  ),
  country_label = c(
    "Indonesia", "Vietnam", "Thailand", "Malaysia", "Philippines", "China", "India",
    "Brazil", "Mexico", "Chile",
    "Japan", "South Korea", "Germany", "France", "Australia", "United States"
  ),
  comparison_group = c(
    rep("Asian peers", 7),
    rep("High-inequality emerging references", 3),
    rep("Advanced-country benchmarks", 6)
  )
)

latest_gini_comparison <- data$comparative_growth %>%
  inner_join(comparison_countries, by = "country_code") %>%
  drop_na(gini) %>%
  group_by(country_code) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    country_label = reorder(country_label, gini),
    is_indonesia = country_code == "IDN",
    label = paste0(round(gini, 1), " (", year, ")")
  )

latest_gini_comparison %>%
  ggplot(aes(country_label, gini, fill = is_indonesia)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = label), hjust = -0.08, size = 3) +
  coord_flip() +
  facet_wrap(~ comparison_group, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c(`FALSE` = palette[["gray"]], `TRUE` = palette[["red"]])) +
  scale_y_continuous(limits = c(0, 62), breaks = seq(0, 60, by = 10)) +
  labs(
    title = "Latest Available Gini Index: Indonesia and Comparator Countries",
    subtitle = "World Bank; labels show Gini value and observation year. Higher Gini means higher measured inequality.",
    x = NULL, y = "Gini index"
  ) +
  chart_theme() +
  theme(legend.position = "none")
```

## Real GDP Per Capita

```{r gdp-per-capita, fig.width=10, fig.height=5.5}
data$gdp %>%
  drop_na(real_gdp_per_capita_lcu) %>%
  ggplot(aes(year, real_gdp_per_capita_lcu)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Real GDP per Capita",
    subtitle = "World Bank API, constant local currency; price effects already removed",
    x = "Year", y = "Constant Indonesian rupiah"
  ) +
  chart_theme()
```

## Average Consumption By Population Group

```{r group-consumption, fig.width=10, fig.height=6}
broad_groups <- c("Bottom 40%", "Middle 40%", "Top 20%")

group_consumption <- data$welfare_distribution %>%
  filter(group %in% broad_groups) %>%
  mutate(group = factor(group, levels = broad_groups)) %>%
  drop_na(avg_welfare_ppp_daily)

group_consumption_labels <- group_consumption %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_welfare = first(avg_welfare_ppp_daily)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (avg_welfare_ppp_daily / first_welfare - 1) * 100,
    growth_label = sprintf("+%.0f%%", growth_pct)
  )

group_consumption %>%
  ggplot(aes(year, avg_welfare_ppp_daily, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = group_consumption_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c(palette[["green"]], palette[["blue"]], palette[["red"]])) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Average Consumption by Population Group",
    subtitle = "End labels show growth from first available year to final year; 2017 PPP USD per person per day",
    x = "Year", y = "Daily consumption", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

## Distribution Shares By Population Group

```{r distribution-shares, fig.width=10, fig.height=5.5}
ineq_long <- data$welfare_distribution %>%
  filter(group %in% broad_groups) %>%
  mutate(
    group = factor(group, levels = broad_groups),
    share = welfare_share * 100
  ) %>%
  drop_na(share)

ineq_change_labels <- ineq_long %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    change_pp = share - first_share,
    change_label = sprintf("%+.1f pp", change_pp)
  )

ineq_long %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = ineq_change_labels,
    aes(label = change_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(palette[["green"]], palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Distribution Shares by Population Group",
    subtitle = "End labels show percentage-point change from first available year to final year",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 70, 5.5, 5.5))
```

## Tail Distribution Shares

```{r tail-distribution-shares, fig.width=10, fig.height=5.5}
tail <- data$welfare_distribution %>%
  filter(group %in% c("Bottom 10%", "Top 10%")) %>%
  mutate(group = factor(group, levels = c("Bottom 10%", "Top 10%")))

tail_shares <- tail %>%
  select(year, group, welfare_share) %>%
  mutate(share = welfare_share * 100) %>%
  drop_na(share)

tail_share_labels <- tail_shares %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (share / first_share - 1) * 100,
    growth_label = sprintf("%+.0f%%", growth_pct)
  )

tail_shares %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = tail_share_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Bottom 10%" = palette[["green"]],
    "Top 10%" = palette[["red"]]
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Tail Distribution Shares",
    subtitle = "End labels show growth from first available year to final year; World Bank PIP consumption shares",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

## Inflation And CPI

```{r inflation-cpi, fig.width=10, fig.height=5.5}
inflation_long <- data$inflation %>%
  select(year, cpi_total, inflation_yoy) %>%
  pivot_longer(
    cols = c(cpi_total, inflation_yoy),
    names_to = "metric",
    values_to = "value"
  ) %>%
  drop_na(value) %>%
  mutate(metric = recode(
    metric,
    cpi_total = "CPI total index",
    inflation_yoy = "Inflation YoY (%)"
  ))

inflation_long %>%
  ggplot(aes(year, value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c(palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Inflation and CPI Coverage Available Now",
    subtitle = "Total CPI and annual inflation from World Bank",
    x = "Year", y = NULL, color = "Metric"
  ) +
  chart_theme()
```

## Nominal Wages, Prices, And Purchasing Power

```{r wage-vs-cpi-index, fig.width=10, fig.height=5.5}
wage_price_index <- data$labor %>%
  select(year, avg_wage) %>%
  left_join(data$inflation %>% select(year, cpi_total), by = "year") %>%
  filter(year >= 2019) %>%
  drop_na(avg_wage, cpi_total) %>%
  mutate(real_wage = avg_wage / cpi_total * 100) %>%
  mutate(
    nominal_wage_index = avg_wage / first(avg_wage) * 100,
    cpi_index = cpi_total / first(cpi_total) * 100,
    real_wage_index = real_wage / first(real_wage) * 100
  ) %>%
  select(year, nominal_wage_index, cpi_index, real_wage_index) %>%
  pivot_longer(
    cols = ends_with("_index"),
    names_to = "metric",
    values_to = "index"
  ) %>%
  mutate(metric = recode(
    metric,
    nominal_wage_index = "Nominal wage",
    cpi_index = "Consumer prices",
    real_wage_index = "Real wage"
  ))

wage_price_index %>%
  ggplot(aes(year, index, color = metric)) +
  geom_hline(yintercept = 100, color = palette[["gray"]], linetype = "dashed") +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = c(
    "Nominal wage" = palette[["blue"]],
    "Consumer prices" = palette[["red"]],
    "Real wage" = palette[["green"]]
  )) +
  scale_x_continuous(breaks = 2019:2024) +
  labs(
    title = "Nominal Wages, Prices, and Purchasing Power",
    subtitle = "Indexed to 2019 = 100; real wage is nominal wage divided by CPI",
    x = "Year", y = "Index, 2019 = 100", color = "Metric"
  ) +
  chart_theme()
```

## Household Price Pressure

```{r household-price-pressure, fig.width=9, fig.height=6}
household_analysis_start_year <- 2020
household_analysis_end_year <- 2023

category_cpi <- data$inflation %>%
  select(year, cpi_food, cpi_transport, cpi_housing, cpi_education, cpi_healthcare, cpi_restaurant) %>%
  pivot_longer(
    cols = starts_with("cpi_"),
    names_to = "category",
    values_to = "index"
  ) %>%
  drop_na(index) %>%
  filter(year >= household_analysis_start_year) %>%
  mutate(category = recode(
    category,
    cpi_food = "Food, beverages, tobacco",
    cpi_transport = "Transport",
    cpi_housing = "Housing, water, electricity, fuel",
    cpi_education = "Education",
    cpi_healthcare = "Health",
    cpi_restaurant = "Restaurants"
  ))

price_pressure <- bind_rows(
  data$labor %>%
    select(year, avg_wage) %>%
    mutate(metric = "Average monthly earnings", value = avg_wage) %>%
    select(year, metric, value),
  data$inflation %>%
    select(year, cpi_total) %>%
    mutate(metric = "Total CPI", value = cpi_total) %>%
    select(year, metric, value),
  category_cpi %>%
    transmute(year, metric = category, value = index)
) %>%
  filter(between(year, household_analysis_start_year, household_analysis_end_year)) %>%
  drop_na(value) %>%
  mutate(metric = factor(
    metric,
    levels = c(
      "Average monthly earnings",
      "Total CPI",
      "Food, beverages, tobacco",
      "Transport",
      "Housing, water, electricity, fuel",
      "Education",
      "Health",
      "Restaurants"
    )
  )) %>%
  group_by(metric) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    base_value = first(value),
    index_start = value / base_value * 100,
    emphasis = if_else(metric %in% c("Average monthly earnings","Total CPI"), "highlight", "muted"),
    line_type = if_else(emphasis == "highlight", "solid", "dashed")
  ) %>%
  ungroup()

price_pressure_growth <- price_pressure %>%
  group_by(metric) %>%
  arrange(year, .by_group = TRUE) %>%
  summarise(
    growth_pct = last(value) / first(value) - 1,
    .groups = "drop"
  )

price_pressure_labels <- price_pressure %>%
  group_by(metric) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(price_pressure_growth, by = "metric") %>%
  mutate(
    label_x = household_analysis_end_year + 0.06,
    label_y = case_when(
      metric == "Transport" ~ 118.6,
      metric == "Food, beverages, tobacco" ~ 115.8,
      metric == "Total CPI" ~ 110.2,
      metric == "Restaurants" ~ 109.2,
      metric == "Average monthly earnings" ~ 108.1,
      metric == "Health" ~ 106.6,
      metric == "Education" ~ 105.7,
      metric == "Housing, water, electricity, fuel" ~ 104.8,
      TRUE ~ index_start
    ),
    label = str_glue("{metric}  {percent(growth_pct, accuracy = 0.1)}")
  )

price_pressure %>%
  ggplot(aes(year, index_start, group = metric)) +
  geom_hline(yintercept = 100, color = palette[["gray"]], linetype = "dashed") +
  geom_line(aes(color = metric, linetype = line_type, alpha = emphasis), linewidth = 1.05) +
  geom_point(aes(color = metric, alpha = emphasis), size = 1.8) +
  geom_text(
    data = price_pressure_labels,
    aes(x = label_x, y = label_y, label = label, color = metric),
    hjust = 0,
    size = 2.9,
    show.legend = FALSE
  ) +
  scale_x_continuous(
    breaks = household_analysis_start_year:household_analysis_end_year,
    limits = c(household_analysis_start_year, household_analysis_end_year + 0.85),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_color_manual(values = c(
    "Average monthly earnings" = palette[["blue"]],
    "Total CPI" = palette[["red"]],
    "Food, beverages, tobacco" = palette[["green"]],
    "Transport" = palette[["gold"]],
    "Housing, water, electricity, fuel" = palette[["purple"]],
    "Education" = "#8A8F98",
    "Health" = "#A0A6AF",
    "Restaurants" = "#747B86"
  )) +
  scale_linetype_manual(values = c(solid = "solid", dashed = "dashed")) +
  scale_alpha_manual(values = c(highlight = 1, muted = 0.55)) +
  scale_y_continuous(
    breaks = seq(95, 125, by = 5),
    limits = c(95, 125)
  ) +
  labs(
    title = "How Prices Move From a Household Starting Point",
    subtitle = str_glue("Each series rebased to its first available value in {household_analysis_start_year}; shown through {household_analysis_end_year}."),
    x = "Year", y = "Index, first available = 100", color = "Series"
  ) +
  coord_cartesian(clip = "off") +
  chart_theme() +
  theme(
    legend.position = "none",
    plot.margin = margin(5.5, 95, 5.5, 5.5)
  )
```

## Average Consumption By Decile

```{r lower-income-welfare-price-pressure, fig.width=9, fig.height=5}
decile_levels <- c(
  "0 - 10%", "11 - 20%", "21 - 30%", "31 - 40%", "41 - 50%",
  "51 - 60%", "61 - 70%", "71 - 80%", "81 - 90%", "91 - 100%"
)

pip_percentiles <- read_csv(
  file.path(project_root, "data", "raw", "worldbank_pip_percentiles_indonesia.csv"),
  show_col_types = FALSE
) %>%
  filter(
    country_code == "IDN",
    reporting_level == "national"
  )

latest_available_consumption_year <- max(pip_percentiles$year, na.rm = TRUE)
consumption_end_year <- min(latest_available_consumption_year, household_analysis_end_year)

ppp_base_year <- 2017

ppp_2017_conversion <- data$gdp %>%
  filter(year == ppp_base_year) %>%
  pull(ppp_private_consumption)

cpi_2017 <- data$inflation %>%
  filter(year == ppp_base_year) %>%
  pull(cpi_total)

latest_cpi <- data$inflation %>%
  filter(year <= consumption_end_year, !is.na(cpi_total)) %>%
  slice_max(year, n = 1, with_ties = FALSE)

latest_cpi_year <- latest_cpi$year
latest_price_conversion <- ppp_2017_conversion * latest_cpi$cpi_total / cpi_2017

decile_consumption <- pip_percentiles %>%
  filter(between(year, household_analysis_start_year, consumption_end_year)) %>%
  mutate(
    percentile = as.integer(percentile),
    decile = case_when(
      between(percentile, 1, 10) ~ "0 - 10%",
      between(percentile, 11, 20) ~ "11 - 20%",
      between(percentile, 21, 30) ~ "21 - 30%",
      between(percentile, 31, 40) ~ "31 - 40%",
      between(percentile, 41, 50) ~ "41 - 50%",
      between(percentile, 51, 60) ~ "51 - 60%",
      between(percentile, 61, 70) ~ "61 - 70%",
      between(percentile, 71, 80) ~ "71 - 80%",
      between(percentile, 81, 90) ~ "81 - 90%",
      between(percentile, 91, 100) ~ "91 - 100%"
    )
  ) %>%
  drop_na(decile, avg_welfare, pop_share) %>%
  group_by(year, decile) %>%
  summarise(
    avg_consumption = weighted.mean(avg_welfare, pop_share),
    .groups = "drop"
  ) %>%
  mutate(decile = factor(decile, levels = decile_levels)) %>%
  arrange(decile, year)

decile_consumption %>%
  group_by(decile) %>%
  arrange(year, .by_group = TRUE) %>%
  summarise(
    consumption_2020 = first(avg_consumption),
    latest_consumption = last(avg_consumption),
    growth = latest_consumption / consumption_2020 - 1,
    .groups = "drop"
  ) %>%
  mutate(
    latest_consumption_rupiah = latest_consumption * latest_price_conversion,
    consumption_label = str_glue("Rp {number(latest_consumption_rupiah, accuracy = 100, big.mark = ',')} / day"),
    label_vjust = if_else(growth >= 0, -0.35, 1.25)
  ) %>%
  ggplot(aes(decile, growth)) +
  geom_hline(yintercept = 0, color = palette[["gray"]], linetype = "dashed") +
  geom_col(fill = palette[["green"]], width = 0.72) +
  geom_text(
    aes(label = consumption_label, vjust = label_vjust),
    size = 3
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0.08, 0.14))
  ) +
  labs(
    title = "Average Consumption Growth by Decile",
    subtitle = str_glue("Bar height shows {household_analysis_start_year}-{consumption_end_year} growth in PIP's constant 2017 PPP consumption welfare."),
    x = "Population decile", y = "Consumption growth",
    caption = str_glue("Labels show latest consumption translated to rupiah/day in {latest_cpi_year} prices: 2017 PPP welfare x 2017 private-consumption PPP x CPI growth since 2017.")
  ) +
  chart_theme() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    legend.position = "none"
  )
```

## Next EDA Tasks

- Validate ILOSTAT average monthly earnings against BPS/Sakernas wage tables if a cleaner national API series is confirmed.
- Extend CPI category history before 2020 and bridge the 2024 BPS 2022=100 base only if the bridge is source-backed.
- Extend cost-of-living charts beyond wholesale rice once fuel/electricity/LPG/rent source IDs are mapped.
- Replace provisional distribution shares with BPS/SUSENAS expenditure-only distribution if available.
- Add informal employment charts once BPS or ILOSTAT source series are confirmed.
