---
title: "Indonesia Inequality Project: EDA - Appendix"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

This appendix runs independently from the main EDA notebook. It loads the same processed data files directly and recreates the shared plotting helpers locally.

```{r setup}
library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

project_root <- if (basename(getwd()) == "notebooks") dirname(getwd()) else getwd()
data_dir <- file.path(project_root, "data", "processed")

palette <- c(
  blue = "#2F6B9A",
  red = "#C44E52",
  green = "#4C956C",
  gold = "#D4A017",
  purple = "#7D5BA6",
  gray = "#6C757D",
  light = "#F6F7F9",
  dark = "#1F2933"
)

chart_theme <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14, color = palette[["dark"]]),
      plot.subtitle = element_text(size = 10, color = palette[["gray"]]),
      axis.title = element_text(size = 10, color = palette[["dark"]]),
      axis.text = element_text(size = 9, color = palette[["dark"]]),
      legend.title = element_text(size = 9, face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

read_processed_csv <- function(filename) {
  read_csv(file.path(data_dir, filename), show_col_types = FALSE) %>%
    rename_with(str_trim)
}
```

## Load Data

```{r load-data}
files <- c(
  gdp = "gdp_indonesia.csv",
  inequality = "inequality_indonesia.csv",
  welfare_distribution = "welfare_distribution_indonesia.csv",
  comparative_growth = "comparative_inequality_growth.csv",
  labor = "labor_indonesia.csv",
  prices = "prices_indonesia.csv"
)

data <- files %>%
  imap(~ read_processed_csv(.x)) %>%
  set_names(names(files))

indonesia_data <- data[names(data) != "comparative_growth"]

tail <- data$welfare_distribution %>%
  filter(group %in% c("Bottom 10%", "Top 10%")) %>%
  mutate(group = factor(group, levels = c("Bottom 10%", "Top 10%")))
```

# Appendix

## Indonesia's Relative Inequality Position

```{r comparative-gini-position, fig.width=10, fig.height=6}
comparison_countries <- tibble(
  country_code = c(
    "IDN", "VNM", "THA", "MYS", "PHL", "CHN", "IND",
    "BRA", "MEX", "CHL",
    "JPN", "KOR", "DEU", "FRA", "AUS", "USA"
  ),
  country_label = c(
    "Indonesia", "Vietnam", "Thailand", "Malaysia", "Philippines", "China", "India",
    "Brazil", "Mexico", "Chile",
    "Japan", "South Korea", "Germany", "France", "Australia", "United States"
  ),
  comparison_group = c(
    rep("Asian peers", 7),
    rep("High-inequality emerging references", 3),
    rep("Advanced-country benchmarks", 6)
  )
)

latest_gini_comparison <- data$comparative_growth %>%
  inner_join(comparison_countries, by = "country_code") %>%
  drop_na(gini) %>%
  group_by(country_code) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  mutate(
    country_label = reorder(country_label, gini),
    is_indonesia = country_code == "IDN",
    label = paste0(round(gini, 1), " (", year, ")")
  )

latest_gini_comparison %>%
  ggplot(aes(country_label, gini, fill = is_indonesia)) +
  geom_col(width = 0.72) +
  geom_text(aes(label = label), hjust = -0.08, size = 3) +
  coord_flip() +
  facet_wrap(~ comparison_group, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c(`FALSE` = palette[["gray"]], `TRUE` = palette[["red"]])) +
  scale_y_continuous(limits = c(0, 62), breaks = seq(0, 60, by = 10)) +
  labs(
    title = "Latest Available Gini Index: Indonesia and Comparator Countries",
    subtitle = "World Bank; labels show Gini value and observation year. Higher Gini means higher measured inequality.",
    x = NULL, y = "Gini index"
  ) +
  chart_theme() +
  theme(legend.position = "none")
```

## Data Validation

```{r data-summary}
data %>%
  imap_dfr(~ tibble(
    dataset = .y,
    rows = nrow(.x),
    first_year = min(.x$year, na.rm = TRUE),
    last_year = max(.x$year, na.rm = TRUE),
    columns = ncol(.x)
  ))
```

```{r data-coverage, fig.width=10, fig.height=5.5}
coverage <- indonesia_data %>%
  imap_dfr(function(df, dataset_name) {
    value_cols <- df %>%
      select(-any_of(c("year", "source"))) %>%
      names()

    df %>%
      rowwise() %>%
      mutate(
        dataset = dataset_name,
        available_fields = sum(!is.na(pick(all_of(value_cols)))),
        total_fields = length(value_cols),
        coverage_rate = if_else(total_fields > 0, available_fields / total_fields, 0)
      ) %>%
      ungroup() %>%
      select(dataset, year, available_fields, total_fields, coverage_rate)
  })

coverage %>%
  ggplot(aes(year, dataset, fill = coverage_rate)) +
  geom_tile(color = "white", linewidth = 0.25) +
  scale_fill_gradient(low = "#F1F3F5", high = palette[["blue"]], labels = percent_format()) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Processed Data Coverage by Dataset and Year",
    subtitle = "Darker cells have more populated fields in the standardized CSV schema",
    x = "Year", y = NULL, fill = "Coverage"
  ) +
  chart_theme()
```

```{r field-availability-table}
availability <- indonesia_data %>%
  imap_dfr(function(df, dataset_name) {
    df %>%
      select(-any_of(c("year", "source"))) %>%
      summarise(across(everything(), ~ sum(!is.na(.x)))) %>%
      pivot_longer(everything(), names_to = "field", values_to = "non_missing_count") %>%
      mutate(
        dataset = dataset_name,
        total_rows = nrow(df),
        status = case_when(
          non_missing_count == 0 ~ "Empty",
          non_missing_count == total_rows ~ "Complete",
          TRUE ~ "Partly missing"
        )
      )
  }) %>%
  mutate(status = factor(status, levels = c("Complete", "Partly missing", "Empty")))

availability %>%
  arrange(dataset, status, field)
```

```{r field-availability-chart, fig.width=10, fig.height=5.5}
availability %>%
  count(dataset, status, name = "field_count") %>%
  ggplot(aes(dataset, field_count, fill = status)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c(
    Complete = palette[["green"]],
    `Partly missing` = palette[["gold"]],
    Empty = palette[["red"]]
  )) +
  labs(
    title = "Field Availability by Dataset",
    subtitle = "Empty fields are mostly waiting for source mapping or API recovery",
    x = NULL, y = "Number of fields", fill = "Status"
  ) +
  chart_theme() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r empty-fields}
availability %>%
  filter(status == "Empty") %>%
  select(dataset, field) %>%
  arrange(dataset, field)
```

```{r combined-dataset}
combined <- data$gdp %>%
  select(-source)

for (name in c("prices", "inequality", "labor")) {
  df <- data[[name]] %>%
    select(-source)

  duplicate_cols <- intersect(names(df), names(combined)) %>%
    setdiff("year")

  df <- df %>%
    select(-any_of(duplicate_cols))

  combined <- combined %>%
    full_join(df, by = "year")
}

combined <- combined %>%
  arrange(year)

combined %>%
  head()
```

## Supporting Charts

```{r gdp-growth, fig.width=10, fig.height=5.5}
data$gdp %>%
  drop_na(gdp_growth) %>%
  ggplot(aes(year, gdp_growth)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = palette[["gray"]]) +
  geom_line(color = palette[["blue"]], linewidth = 1.1) +
  geom_point(color = palette[["blue"]], size = 1.8) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "GDP Growth, Year over Year",
    subtitle = "World Bank API, annual percent",
    x = "Year", y = "GDP growth (%)"
  ) +
  chart_theme()
```
## Average Consumption By Population Group

```{r group-consumption, fig.width=10, fig.height=6}
broad_groups <- c("Bottom 40%", "Middle 40%", "Top 20%")

group_consumption <- data$welfare_distribution %>%
  filter(group %in% broad_groups) %>%
  mutate(group = factor(group, levels = broad_groups)) %>%
  drop_na(avg_welfare_ppp_daily)

group_consumption_labels <- group_consumption %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_welfare = first(avg_welfare_ppp_daily)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (avg_welfare_ppp_daily / first_welfare - 1) * 100,
    growth_label = sprintf("+%.0f%%", growth_pct)
  )

group_consumption %>%
  ggplot(aes(year, avg_welfare_ppp_daily, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = group_consumption_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c(palette[["green"]], palette[["blue"]], palette[["red"]])) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Average Consumption by Population Group",
    subtitle = "End labels show growth from first available year to final year; 2017 PPP USD per person per day",
    x = "Year", y = "Daily consumption", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

## Distribution Shares By Population Group

```{r distribution-shares, fig.width=10, fig.height=5.5}
ineq_long <- data$welfare_distribution %>%
  filter(group %in% broad_groups) %>%
  mutate(
    group = factor(group, levels = broad_groups),
    share = welfare_share * 100
  ) %>%
  drop_na(share)

ineq_change_labels <- ineq_long %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    change_pp = share - first_share,
    change_label = sprintf("%+.1f pp", change_pp)
  )

ineq_long %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = ineq_change_labels,
    aes(label = change_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(palette[["green"]], palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Distribution Shares by Population Group",
    subtitle = "End labels show percentage-point change from first available year to final year",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 70, 5.5, 5.5))
```

```{r tail-absolute-consumption, fig.width=10, fig.height=5.5}
tail_consumption <- tail %>%
  select(year, group, avg_welfare_ppp_daily) %>%
  drop_na(avg_welfare_ppp_daily)

tail_consumption_labels <- tail_consumption %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_welfare = first(avg_welfare_ppp_daily)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (avg_welfare_ppp_daily / first_welfare - 1) * 100,
    growth_label = sprintf("+%.0f%%", growth_pct)
  )

tail_consumption %>%
  ggplot(aes(year, avg_welfare_ppp_daily, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = tail_consumption_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Bottom 10%" = palette[["green"]],
    "Top 10%" = palette[["red"]]
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Tail Average Consumption",
    subtitle = "End labels show growth from first available year to final year; 2017 PPP USD per person per day",
    x = "Year", y = "Daily consumption", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

```{r extreme-tail-distribution-shares, fig.width=10, fig.height=5.5}
extreme_tail <- data$welfare_distribution %>%
  filter(group %in% c("Bottom 1%", "Top 1%")) %>%
  mutate(
    group = factor(group, levels = c("Bottom 1%", "Top 1%")),
    share = welfare_share * 100
  ) %>%
  drop_na(share)

extreme_tail_share_labels <- extreme_tail %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (share / first_share - 1) * 100,
    growth_label = sprintf("%+.0f%%", growth_pct)
  )

extreme_tail %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = extreme_tail_share_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Bottom 1%" = palette[["green"]],
    "Top 1%" = palette[["red"]]
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Extreme Tail Distribution Shares",
    subtitle = "End labels show growth from first available year to final year; World Bank PIP consumption shares",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```
## Inflation And CPI

```{r inflation-cpi, fig.width=10, fig.height=5.5}
inflation_long <- data$prices %>%
  select(year, cpi_total, inflation_yoy) %>%
  pivot_longer(
    cols = c(cpi_total, inflation_yoy),
    names_to = "metric",
    values_to = "value"
  ) %>%
  drop_na(value) %>%
  mutate(metric = recode(
    metric,
    cpi_total = "CPI total index",
    inflation_yoy = "Inflation YoY (%)"
  ))

inflation_long %>%
  ggplot(aes(year, value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c(palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Inflation and CPI Coverage Available Now",
    subtitle = "Total CPI and annual inflation from World Bank",
    x = "Year", y = NULL, color = "Metric"
  ) +
  chart_theme()
```

```{r gini-index, fig.width=10, fig.height=5.5}
data$inequality %>%
  drop_na(gini_ratio) %>%
  ggplot(aes(year, gini_ratio)) +
  geom_line(color = palette[["purple"]], linewidth = 1.1) +
  geom_point(color = palette[["purple"]], size = 1.8) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Gini Index",
    subtitle = "World Bank API, Indonesia",
    x = "Year", y = "Gini index"
  ) +
  chart_theme()
```

```{r comparative-inequality-growth, fig.width=10, fig.height=5.8}
comparative_long <- data$comparative_growth %>%
  filter(income_level %in% c("Lower middle income", "Upper middle income")) %>%
  select(
    country_code, country_name, region, income_level, year, gini,
    future_5y_growth_pct, future_10y_growth_pct
  ) %>%
  pivot_longer(
    cols = c(future_5y_growth_pct, future_10y_growth_pct),
    names_to = "horizon",
    values_to = "future_growth_pct"
  ) %>%
  mutate(
    horizon = recode(
      horizon,
      future_5y_growth_pct = "Next 5 years",
      future_10y_growth_pct = "Next 10 years"
    ),
    focus = if_else(country_code == "IDN", "Indonesia", "Other middle-income countries")
  ) %>%
  drop_na(gini, future_growth_pct)

comparative_long %>%
  ggplot(aes(gini, future_growth_pct)) +
  geom_hline(yintercept = 0, color = palette[["gray"]], linetype = "dashed") +
  geom_point(
    data = comparative_long %>% filter(focus != "Indonesia"),
    color = palette[["gray"]],
    alpha = 0.28,
    size = 1.7
  ) +
  geom_smooth(
    data = comparative_long %>% filter(focus != "Indonesia"),
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = palette[["purple"]],
    linewidth = 1
  ) +
  geom_point(
    data = comparative_long %>% filter(focus == "Indonesia"),
    color = palette[["red"]],
    size = 2.4
  ) +
  facet_wrap(~ horizon, ncol = 2) +
  scale_x_continuous(limits = c(20, 65), breaks = seq(20, 65, by = 10)) +
  labs(
    title = "Inequality Today and Future Growth",
    subtitle = "Middle-income country-years; red points highlight Indonesia. Trend line is descriptive, not causal.",
    x = "Gini index in year t",
    y = "Annualized future real GDP per-capita growth (%)"
  ) +
  chart_theme()
```

```{r rice-price, fig.width=10, fig.height=5.5}
data$prices %>%
  drop_na(rice_price) %>%
  ggplot(aes(year, rice_price)) +
  geom_line(color = palette[["green"]], linewidth = 1.1) +
  geom_point(color = palette[["green"]], size = 1.8) +
  scale_x_continuous(breaks = seq(2010, 2024, by = 2)) +
  scale_y_continuous(labels = label_number(big.mark = ",")) +
  labs(
    title = "Wholesale Rice Price From BPS",
    subtitle = "BPS WebAPI variable 295; rupiah per kilogram, annual average",
    x = "Year", y = "Rupiah per kg"
  ) +
  chart_theme()
```

```{r wage-price-summary}
data$labor %>%
  select(year, avg_wage, real_wage_growth) %>%
  left_join(data$prices %>% select(year, inflation_yoy), by = "year") %>%
  filter(year >= 2019, year <= 2023) %>%
  mutate(
    nominal_wage_growth = (avg_wage / lag(avg_wage) - 1) * 100,
    real_wage_growth = round(real_wage_growth, 2),
    nominal_wage_growth = round(nominal_wage_growth, 2),
    inflation_yoy = round(inflation_yoy, 2)
  ) %>%
  select(year, avg_wage, nominal_wage_growth, inflation_yoy, real_wage_growth)
```

```{r cost-of-living-missingness, fig.width=10, fig.height=5.5}
data$prices %>%
  select(-any_of(c("year", "source"))) %>%
  summarise(across(everything(), ~ mean(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "field", values_to = "missing_rate") %>%
  ggplot(aes(missing_rate, field)) +
  geom_col(fill = palette[["gold"]]) +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Cost-of-Living Field Missingness",
    subtitle = "Total CPI and BPS wholesale rice price are populated; remaining fields need source mapping",
    x = "Share of rows missing", y = NULL
  ) +
  chart_theme()
```
## Tail Distribution Shares

```{r tail-distribution-shares, fig.width=10, fig.height=5.5}
tail <- data$welfare_distribution %>%
  filter(group %in% c("Bottom 10%", "Top 10%")) %>%
  mutate(group = factor(group, levels = c("Bottom 10%", "Top 10%")))

tail_shares <- tail %>%
  select(year, group, welfare_share) %>%
  mutate(share = welfare_share * 100) %>%
  drop_na(share)

tail_share_labels <- tail_shares %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (share / first_share - 1) * 100,
    growth_label = sprintf("%+.0f%%", growth_pct)
  )

tail_shares %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = tail_share_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Bottom 10%" = palette[["green"]],
    "Top 10%" = palette[["red"]]
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Tail Distribution Shares",
    subtitle = "End labels show growth from first available year to final year; World Bank PIP consumption shares",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

## Decile Consumption Detail

```{r appendix-decile-consumption-prep}
decile_levels <- c(
  "0 - 10%", "11 - 20%", "21 - 30%", "31 - 40%", "41 - 50%",
  "51 - 60%", "61 - 70%", "71 - 80%", "81 - 90%", "91 - 100%"
)

pip_percentiles <- read_csv(
  file.path(project_root, "data", "raw", "worldbank_pip_percentiles_indonesia.csv"),
  show_col_types = FALSE
) %>%
  filter(
    country_code == "IDN",
    reporting_level == "national"
  )

latest_available_consumption_year <- max(pip_percentiles$year, na.rm = TRUE)
long_horizon_start_year <- 2000
long_horizon_end_year <- min(latest_available_consumption_year, 2024)

ppp_base_year <- 2017

ppp_2017_conversion <- data$gdp %>%
  filter(year == ppp_base_year) %>%
  pull(ppp_private_consumption)

cpi_2017 <- data$prices %>%
  filter(year == ppp_base_year) %>%
  pull(cpi_total)

long_horizon_cpi <- data$prices %>%
  filter(year <= long_horizon_end_year, !is.na(cpi_total)) %>%
  slice_max(year, n = 1, with_ties = FALSE)

long_horizon_cpi_year <- long_horizon_cpi$year
long_horizon_price_conversion <- ppp_2017_conversion * long_horizon_cpi$cpi_total / cpi_2017

decile_consumption_long <- pip_percentiles %>%
  filter(between(year, long_horizon_start_year, long_horizon_end_year)) %>%
  mutate(
    percentile = as.integer(percentile),
    decile = case_when(
      between(percentile, 1, 10) ~ "0 - 10%",
      between(percentile, 11, 20) ~ "11 - 20%",
      between(percentile, 21, 30) ~ "21 - 30%",
      between(percentile, 31, 40) ~ "31 - 40%",
      between(percentile, 41, 50) ~ "41 - 50%",
      between(percentile, 51, 60) ~ "51 - 60%",
      between(percentile, 61, 70) ~ "61 - 70%",
      between(percentile, 71, 80) ~ "71 - 80%",
      between(percentile, 81, 90) ~ "81 - 90%",
      between(percentile, 91, 100) ~ "91 - 100%"
    )
  ) %>%
  drop_na(decile, avg_welfare, pop_share) %>%
  group_by(year, decile) %>%
  summarise(
    avg_consumption = weighted.mean(avg_welfare, pop_share),
    .groups = "drop"
  ) %>%
  mutate(decile = factor(decile, levels = decile_levels)) %>%
  arrange(decile, year)

cpi_index <- data$prices %>%
  filter(between(year, long_horizon_start_year, long_horizon_end_year)) %>%
  arrange(year) %>%
  mutate(cpi_index_2000 = cpi_total / first(cpi_total) * 100) %>%
  select(year, cpi_index_2000)

decile_consumption_index <- decile_consumption_long %>%
  group_by(decile) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(real_consumption_index_2000 = avg_consumption / first(avg_consumption) * 100) %>%
  ungroup() %>%
  left_join(cpi_index, by = "year") %>%
  transmute(
    year,
    series = as.character(decile),
    index_2000 = real_consumption_index_2000 * cpi_index_2000 / 100
  )

index_line_palette <- set_names(hcl.colors(length(decile_levels), palette = "Dark 3"), decile_levels)
```

```{r decile-consumption-long-horizon, fig.width=9, fig.height=5}
decile_consumption_long %>%
  group_by(decile) %>%
  arrange(year, .by_group = TRUE) %>%
  summarise(
    consumption_2000 = first(avg_consumption),
    latest_consumption = last(avg_consumption),
    growth = latest_consumption / consumption_2000 - 1,
    .groups = "drop"
  ) %>%
  mutate(
    latest_consumption_rupiah = latest_consumption * long_horizon_price_conversion,
    consumption_label = str_glue("Rp {number(latest_consumption_rupiah, accuracy = 100, big.mark = ',')} / day"),
    label_vjust = if_else(growth >= 0, -0.35, 1.25)
  ) %>%
  ggplot(aes(decile, growth)) +
  geom_hline(yintercept = 0, color = palette[["gray"]], linetype = "dashed") +
  geom_col(fill = palette[["green"]], width = 0.72) +
  geom_text(
    aes(label = consumption_label, vjust = label_vjust),
    size = 3
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0.08, 0.14))
  ) +
  labs(
    title = "Long-Run Average Consumption Growth by Decile",
    subtitle = str_glue("Bar height shows {long_horizon_start_year}-{long_horizon_end_year} growth in PIP's constant 2017 PPP consumption welfare."),
    x = "Population decile", y = "Consumption growth",
    caption = str_glue("Labels show {long_horizon_end_year} consumption translated to rupiah/day in {long_horizon_cpi_year} prices: 2017 PPP welfare x 2017 private-consumption PPP x CPI growth since 2017.")
  ) +
  chart_theme() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    legend.position = "none"
  )
```

```{r decile-consumption-yoy-growth, fig.width=10, fig.height=6}
decile_yoy_start_year <- 2020
decile_yoy_end_year <- min(long_horizon_end_year, 2024)

decile_consumption_yoy <- decile_consumption_index %>%
  mutate(series = factor(series, levels = decile_levels)) %>%
  group_by(series) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(yoy_growth = index_2000 / lag(index_2000) - 1) %>%
  ungroup() %>%
  filter(between(year, decile_yoy_start_year, decile_yoy_end_year)) %>%
  drop_na(yoy_growth)

decile_consumption_yoy_labels <- decile_consumption_yoy %>%
  group_by(series) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(yoy_growth) %>%
  mutate(
    label_x = decile_yoy_end_year + 0.08,
    label_y = case_when(
      series == "0 - 10%" ~ 0.072,
      series == "11 - 20%" ~ 0.064,
      series == "21 - 30%" ~ 0.059,
      series == "31 - 40%" ~ 0.054,
      series == "41 - 50%" ~ 0.049,
      series == "51 - 60%" ~ 0.044,
      series == "61 - 70%" ~ 0.039,
      series == "71 - 80%" ~ 0.034,
      series == "81 - 90%" ~ 0.029,
      series == "91 - 100%" ~ -0.005,
      TRUE ~ yoy_growth
    ),
    label = str_replace_all(as.character(series), " - ", "-")
  )

decile_consumption_yoy %>%
  ggplot(aes(year, yoy_growth, color = series)) +
  geom_hline(yintercept = 0, color = palette[["gray"]], linetype = "dashed") +
  geom_line(linewidth = 0.95) +
  geom_point(size = 1.5) +
  geom_text(
    data = decile_consumption_yoy_labels,
    aes(x = label_x, y = label_y, label = label, color = series),
    hjust = 0,
    size = 2.8,
    show.legend = FALSE
  ) +
  scale_x_continuous(
    breaks = decile_yoy_start_year:decile_yoy_end_year,
    limits = c(decile_yoy_start_year, decile_yoy_end_year + 0.8),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_color_manual(values = index_line_palette) +
  labs(
    title = "YoY Growth in Nominal Average Consumption by Decile",
    subtitle = str_glue("Year-over-year change in implied nominal consumption indexes; focused on {decile_yoy_start_year}-{decile_yoy_end_year}."),
    x = "Year", y = "YoY nominal growth", color = "Decile"
  ) +
  coord_cartesian(clip = "off") +
  chart_theme() +
  theme(
    legend.position = "none",
    plot.margin = margin(5.5, 78, 5.5, 5.5)
  )
```
