---
title: "Indonesia Inequality Project: Initial EDA"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

This Quarto notebook mirrors `eda_indonesia_inequality.ipynb` using the tidyverse paradigm, `%>%` pipes, and `ggplot2`.

Scope for this notebook:

- inspect processed CSV coverage
- visualize available World Bank-backed fields
- make BPS-related gaps visible
- avoid modeling, regressions, causal claims, or final narrative conclusions

```{r setup}
library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

project_root <- if (basename(getwd()) == "notebooks") dirname(getwd()) else getwd()
data_dir <- file.path(project_root, "data", "processed")

palette <- c(
  blue = "#2F6B9A",
  red = "#C44E52",
  green = "#4C956C",
  gold = "#D4A017",
  purple = "#7D5BA6",
  gray = "#6C757D",
  light = "#F6F7F9",
  dark = "#1F2933"
)

chart_theme <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14, color = palette[["dark"]]),
      plot.subtitle = element_text(size = 10, color = palette[["gray"]]),
      axis.title = element_text(size = 10, color = palette[["dark"]]),
      axis.text = element_text(size = 9, color = palette[["dark"]]),
      legend.title = element_text(size = 9, face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

read_processed_csv <- function(filename) {
  read_csv(file.path(data_dir, filename), show_col_types = FALSE) %>%
    rename_with(str_trim)
}
```

## Load Processed Data

Each CSV is loaded from `data/processed`. The pipeline currently gives annual rows for 2000-2024 after Phase 1 is run locally. Some cells are intentionally blank where BPS-specific sources are not yet available.

```{r load-data}
files <- c(
  gdp = "gdp_indonesia.csv",
  inflation = "inflation_indonesia.csv",
  inequality = "inequality_indonesia.csv",
  welfare_distribution = "welfare_distribution_indonesia.csv",
  comparative_growth = "comparative_inequality_growth.csv",
  poverty = "poverty_indonesia.csv",
  labor = "labor_indonesia.csv",
  cost_of_living = "cost_of_living_indonesia.csv"
)

data <- files %>%
  imap(~ read_processed_csv(.x)) %>%
  set_names(names(files))

indonesia_data <- data[names(data) != "comparative_growth"]

data %>%
  imap_dfr(~ tibble(
    dataset = .y,
    rows = nrow(.x),
    first_year = min(.x$year, na.rm = TRUE),
    last_year = max(.x$year, na.rm = TRUE),
    columns = ncol(.x)
  ))
```

## Data Coverage

This view checks which dataset-year combinations contain any non-source values beyond `year`. It is a quick QA step before looking at individual indicators.

```{r data-coverage, fig.width=10, fig.height=5.5}
coverage <- indonesia_data %>%
  imap_dfr(function(df, dataset_name) {
    value_cols <- df %>%
      select(-any_of(c("year", "source"))) %>%
      names()

    df %>%
      rowwise() %>%
      mutate(
        dataset = dataset_name,
        available_fields = sum(!is.na(pick(all_of(value_cols)))),
        total_fields = length(value_cols),
        coverage_rate = if_else(total_fields > 0, available_fields / total_fields, 0)
      ) %>%
      ungroup() %>%
      select(dataset, year, available_fields, total_fields, coverage_rate)
  })

coverage %>%
  ggplot(aes(year, dataset, fill = coverage_rate)) +
  geom_tile(color = "white", linewidth = 0.25) +
  scale_fill_gradient(low = "#F1F3F5", high = palette[["blue"]], labels = percent_format()) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Processed Data Coverage by Dataset and Year",
    subtitle = "Darker cells have more populated fields in the standardized CSV schema",
    x = "Year", y = NULL, fill = "Coverage"
  ) +
  chart_theme()
```

## Field Availability

Instead of plotting every column name, this section groups fields into three plain statuses:

- **Complete**: every year has a value
- **Partly missing**: some years have values, some years are blank
- **Empty**: no years have values yet

The empty fields are mostly BPS-dependent fields. They are present in the schema so the pipeline has a stable destination once BPS/API/manual sources are added.

```{r field-availability-table}
availability <- indonesia_data %>%
  imap_dfr(function(df, dataset_name) {
    df %>%
      select(-any_of(c("year", "source"))) %>%
      summarise(across(everything(), ~ sum(!is.na(.x)))) %>%
      pivot_longer(everything(), names_to = "field", values_to = "non_missing_count") %>%
      mutate(
        dataset = dataset_name,
        total_rows = nrow(df),
        status = case_when(
          non_missing_count == 0 ~ "Empty",
          non_missing_count == total_rows ~ "Complete",
          TRUE ~ "Partly missing"
        )
      )
  }) %>%
  mutate(status = factor(status, levels = c("Complete", "Partly missing", "Empty")))

availability %>%
  arrange(dataset, status, field)
```

```{r field-availability-chart, fig.width=10, fig.height=5.5}
availability %>%
  count(dataset, status, name = "field_count") %>%
  ggplot(aes(dataset, field_count, fill = status)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c(
    Complete = palette[["green"]],
    `Partly missing` = palette[["gold"]],
    Empty = palette[["red"]]
  )) +
  labs(
    title = "Field Availability by Dataset",
    subtitle = "Empty fields are mostly waiting for BPS source mapping or API recovery",
    x = NULL, y = "Number of fields", fill = "Status"
  ) +
  chart_theme() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

### Empty Fields to Source Later

These are the fields that currently have no values in any year. They are not analysis-ready yet.

```{r empty-fields}
availability %>%
  filter(status == "Empty") %>%
  select(dataset, field) %>%
  arrange(dataset, field)
```

## Real GDP and Per-Capita Output

The main growth story uses real GDP in local currency. This removes domestic price effects and avoids making the first chart depend on US-dollar exchange-rate movement.

```{r gdp-per-capita, fig.width=10, fig.height=5.5}
data$gdp %>%
  drop_na(real_gdp_per_capita_lcu) %>%
  ggplot(aes(year, real_gdp_per_capita_lcu)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Real GDP per Capita",
    subtitle = "World Bank API, constant local currency; price effects already removed",
    x = "Year", y = "Constant Indonesian rupiah"
  ) +
  chart_theme()
```

```{r gdp-growth, fig.width=10, fig.height=5.5}
data$gdp %>%
  drop_na(gdp_growth) %>%
  ggplot(aes(year, gdp_growth)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = palette[["gray"]]) +
  geom_line(color = palette[["blue"]], linewidth = 1.1) +
  geom_point(color = palette[["blue"]], size = 1.8) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "GDP Growth, Year over Year",
    subtitle = "World Bank API, annual percent",
    x = "Year", y = "GDP growth (%)"
  ) +
  chart_theme()
```

## Inflation and CPI

The total CPI and annual inflation fields are available from World Bank. Category CPI fields are present in the schema but remain blank until BPS source tables are mapped.

```{r inflation-cpi, fig.width=10, fig.height=5.5}
inflation_long <- data$inflation %>%
  select(year, cpi_total, inflation_yoy) %>%
  pivot_longer(
    cols = c(cpi_total, inflation_yoy),
    names_to = "metric",
    values_to = "value"
  ) %>%
  drop_na(value) %>%
  mutate(metric = recode(
    metric,
    cpi_total = "CPI total index",
    inflation_yoy = "Inflation YoY (%)"
  ))

inflation_long %>%
  ggplot(aes(year, value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c(palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Inflation and CPI Coverage Available Now",
    subtitle = "Category CPI fields require BPS follow-up",
    x = "Year", y = NULL, color = "Metric"
  ) +
  chart_theme()
```

## Inequality and Distribution Shares

The pipeline uses World Bank PIP percentile data for absolute group-level welfare and World Bank inequality indicators for Gini. For Indonesia, the current PIP welfare type is consumption, so group-level welfare can be read as average consumption per person per day in 2017 PPP dollars.

```{r group-consumption, fig.width=10, fig.height=6}
broad_groups <- c("Bottom 40%", "Middle 40%", "Top 20%")

group_consumption <- data$welfare_distribution %>%
  filter(group %in% broad_groups) %>%
  mutate(group = factor(group, levels = broad_groups)) %>%
  drop_na(avg_welfare_ppp_daily)

group_consumption_labels <- group_consumption %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_welfare = first(avg_welfare_ppp_daily)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (avg_welfare_ppp_daily / first_welfare - 1) * 100,
    growth_label = sprintf("+%.0f%%", growth_pct)
  )

group_consumption %>%
  ggplot(aes(year, avg_welfare_ppp_daily, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = group_consumption_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 3) +
  scale_color_manual(values = c(palette[["green"]], palette[["blue"]], palette[["red"]])) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Average Consumption by Population Group",
    subtitle = "End labels show growth from first available year to final year; 2017 PPP USD per person per day",
    x = "Year", y = "Daily consumption", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

The broad bottom 40% / middle 40% / top 20% share view answers a different question: whether each group receives a larger or smaller slice of total consumption. The added bottom 10% and top 10% view checks whether distribution changes are more extreme at the ends.

```{r distribution-shares, fig.width=10, fig.height=5.5}
ineq_long <- data$welfare_distribution %>%
  filter(group %in% broad_groups) %>%
  mutate(
    group = factor(group, levels = broad_groups),
    share = welfare_share * 100
  ) %>%
  drop_na(share)

ineq_change_labels <- ineq_long %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    change_pp = share - first_share,
    change_label = sprintf("%+.1f pp", change_pp)
  )

ineq_long %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = ineq_change_labels,
    aes(label = change_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(palette[["green"]], palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Distribution Shares by Population Group",
    subtitle = "End labels show percentage-point change from first available year to final year",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 70, 5.5, 5.5))
```

### Tail Distribution Shares

This view compares the bottom 10% and top 10% in two ways: absolute average consumption and share of total consumption. The absolute chart answers whether each tail improved in level terms. The share chart answers whether each tail received a larger or smaller slice of total consumption.

```{r tail-absolute-consumption, fig.width=10, fig.height=5.5}
tail <- data$welfare_distribution %>%
  filter(group %in% c("Bottom 10%", "Top 10%")) %>%
  mutate(group = factor(group, levels = c("Bottom 10%", "Top 10%")))

tail_consumption <- tail %>%
  select(year, group, avg_welfare_ppp_daily) %>%
  drop_na(avg_welfare_ppp_daily)

tail_consumption_labels <- tail_consumption %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_welfare = first(avg_welfare_ppp_daily)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (avg_welfare_ppp_daily / first_welfare - 1) * 100,
    growth_label = sprintf("+%.0f%%", growth_pct)
  )

tail_consumption %>%
  ggplot(aes(year, avg_welfare_ppp_daily, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = tail_consumption_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Bottom 10%" = palette[["green"]],
    "Top 10%" = palette[["red"]]
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Tail Average Consumption",
    subtitle = "End labels show growth from first available year to final year; 2017 PPP USD per person per day",
    x = "Year", y = "Daily consumption", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

```{r tail-distribution-shares, fig.width=10, fig.height=5.5}
tail_shares <- tail %>%
  select(year, group, welfare_share) %>%
  mutate(
    share = welfare_share * 100
  ) %>%
  drop_na(share)

tail_share_labels <- tail_shares %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    growth_pct = (share / first_share - 1) * 100,
    growth_label = sprintf("%+.0f%%", growth_pct)
  )

tail_shares %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = tail_share_labels,
    aes(label = growth_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  facet_wrap(~ group, scales = "free_y", ncol = 2) +
  scale_color_manual(values = c(
    "Bottom 10%" = palette[["green"]],
    "Top 10%" = palette[["red"]]
  )) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Tail Distribution Shares",
    subtitle = "End labels show growth from first available year to final year; World Bank PIP consumption shares",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

```{r gini-index, fig.width=10, fig.height=5.5}
data$inequality %>%
  drop_na(gini_ratio) %>%
  ggplot(aes(year, gini_ratio)) +
  geom_line(color = palette[["purple"]], linewidth = 1.1) +
  geom_point(color = palette[["purple"]], size = 1.8) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Gini Index",
    subtitle = "World Bank API, Indonesia",
    x = "Year", y = "Gini index"
  ) +
  chart_theme()
```

## Comparative Inequality and Future Growth

This section treats the inequality-growth question comparatively, not as Indonesia-only causal proof. It asks whether country-years with higher inequality tend to be followed by weaker real GDP-per-capita growth.

The chart uses middle-income country observations only, because Indonesia is currently an upper-middle-income country and the comparison is more meaningful among countries at broadly similar development stages.

```{r comparative-inequality-growth, fig.width=10, fig.height=5.8}
comparative_long <- data$comparative_growth %>%
  filter(income_level %in% c("Lower middle income", "Upper middle income")) %>%
  select(
    country_code, country_name, region, income_level, year, gini,
    future_5y_growth_pct, future_10y_growth_pct
  ) %>%
  pivot_longer(
    cols = c(future_5y_growth_pct, future_10y_growth_pct),
    names_to = "horizon",
    values_to = "future_growth_pct"
  ) %>%
  mutate(
    horizon = recode(
      horizon,
      future_5y_growth_pct = "Next 5 years",
      future_10y_growth_pct = "Next 10 years"
    ),
    focus = if_else(country_code == "IDN", "Indonesia", "Other middle-income countries")
  ) %>%
  drop_na(gini, future_growth_pct)

indonesia_labels <- comparative_long %>%
  filter(country_code == "IDN", year %in% c(2000, 2010, 2014))

comparative_long %>%
  ggplot(aes(gini, future_growth_pct)) +
  geom_hline(yintercept = 0, color = palette[["gray"]], linetype = "dashed") +
  geom_point(
    data = comparative_long %>% filter(focus != "Indonesia"),
    color = palette[["gray"]],
    alpha = 0.28,
    size = 1.7
  ) +
  geom_smooth(
    data = comparative_long %>% filter(focus != "Indonesia"),
    method = "lm",
    formula = y ~ x,
    se = FALSE,
    color = palette[["purple"]],
    linewidth = 1
  ) +
  geom_point(
    data = comparative_long %>% filter(focus == "Indonesia"),
    color = palette[["red"]],
    size = 2.4
  ) +
  geom_text(
    data = indonesia_labels,
    aes(label = year),
    color = palette[["red"]],
    nudge_x = 0.8,
    hjust = 0,
    size = 3,
    show.legend = FALSE
  ) +
  facet_wrap(~ horizon, ncol = 2) +
  scale_x_continuous(limits = c(20, 65), breaks = seq(20, 65, by = 10)) +
  labs(
    title = "Inequality Today and Future Growth",
    subtitle = "Middle-income country-years; red points highlight Indonesia. Trend line is descriptive, not causal.",
    x = "Gini index in year t",
    y = "Annualized future real GDP per-capita growth (%)"
  ) +
  chart_theme()
```

```{r comparative-inequality-growth-table}
comparative_long %>%
  filter(horizon == "Next 10 years") %>%
  mutate(gini_band = cut(gini, breaks = c(20, 30, 40, 50, 60, 70), right = FALSE)) %>%
  group_by(gini_band) %>%
  summarise(
    observations = n(),
    median_future_growth = median(future_growth_pct, na.rm = TRUE),
    p25_future_growth = quantile(future_growth_pct, 0.25, na.rm = TRUE),
    p75_future_growth = quantile(future_growth_pct, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(across(ends_with("growth"), ~ round(.x, 2))) %>%
  arrange(gini_band)
```

Interpret this cautiously. The chart can support a comparative risk statement: higher inequality is often discussed as a possible drag on future growth, but this simple view does not prove causality. Country structure, education, investment, institutions, global shocks, and commodity cycles also matter.

## Poverty

Poverty fields are annual World Bank indicators: national poverty line headcount and international extreme poverty headcount.

```{r poverty-indicators, fig.width=10, fig.height=5.5}
poverty_long <- data$poverty %>%
  select(year, poverty_rate, extreme_poverty_rate) %>%
  pivot_longer(
    cols = c(poverty_rate, extreme_poverty_rate),
    names_to = "metric",
    values_to = "value"
  ) %>%
  drop_na(value) %>%
  mutate(metric = recode(
    metric,
    poverty_rate = "National poverty rate",
    extreme_poverty_rate = "Extreme poverty rate"
  ))

poverty_long %>%
  ggplot(aes(year, value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  scale_color_manual(values = c(palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Poverty Indicators",
    subtitle = "World Bank API, annual percent of population",
    x = "Year", y = "Population share (%)", color = "Metric"
  ) +
  chart_theme()
```

## Labor Market

The available labor fields cover total unemployment, youth unemployment, and vulnerable employment. Wage and informal employment fields remain BPS/ILOSTAT follow-up items.

```{r labor-market, fig.width=10, fig.height=5.5}
labor_long <- data$labor %>%
  select(year, unemployment_rate, youth_unemployment, vulnerable_employment_share) %>%
  pivot_longer(
    cols = c(unemployment_rate, youth_unemployment, vulnerable_employment_share),
    names_to = "metric",
    values_to = "value"
  ) %>%
  drop_na(value) %>%
  mutate(metric = recode(
    metric,
    unemployment_rate = "Unemployment",
    youth_unemployment = "Youth unemployment",
    vulnerable_employment_share = "Vulnerable employment"
  ))

labor_long %>%
  ggplot(aes(year, value, color = metric)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_color_manual(values = c(palette[["blue"]], palette[["red"]], palette[["green"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5)) +
  labs(
    title = "Labor Market Indicators Available Now",
    subtitle = "Wages and informal employment are not yet sourced",
    x = "Year", y = "Share (%)", color = "Metric"
  ) +
  chart_theme()
```

## Cost of Living Placeholder

The cost-of-living schema exists, but only total CPI is currently populated. This chart is mostly a reminder that this dataset needs BPS commodity/category sources before substantive EDA.

```{r cost-of-living-missingness, fig.width=10, fig.height=5.5}
data$cost_of_living %>%
  select(-any_of(c("year", "source"))) %>%
  summarise(across(everything(), ~ mean(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "field", values_to = "missing_rate") %>%
  ggplot(aes(missing_rate, field)) +
  geom_col(fill = palette[["gold"]]) +
  scale_x_continuous(labels = percent_format()) +
  labs(
    title = "Cost-of-Living Field Missingness",
    subtitle = "Only total CPI is populated in the current ingestion pass",
    x = "Share of rows missing", y = NULL
  ) +
  chart_theme()
```

## Starter Combined Dataset

This final section creates a wide analytical table by joining all currently available annual fields. It is useful for future EDA, but should not be interpreted as a finalized analysis dataset until BPS gaps are addressed.

```{r combined-dataset}
combined <- data$gdp %>%
  select(-source)

for (name in c("inflation", "inequality", "poverty", "labor", "cost_of_living")) {
  df <- data[[name]] %>%
    select(-source)

  duplicate_cols <- intersect(names(df), names(combined)) %>%
    setdiff("year")

  df <- df %>%
    select(-any_of(duplicate_cols))

  combined <- combined %>%
    full_join(df, by = "year")
}

combined <- combined %>%
  arrange(year)

combined %>%
  head()
```

```{r combined-summary}
combined %>%
  summarise(across(everything(), list(
    count = ~ sum(!is.na(.x)),
    mean = ~ mean(.x, na.rm = TRUE),
    sd = ~ sd(.x, na.rm = TRUE),
    min = ~ min(.x, na.rm = TRUE),
    max = ~ max(.x, na.rm = TRUE)
  ))) %>%
  pivot_longer(
    everything(),
    names_to = c("field", ".value"),
    names_pattern = "(.+)_(count|mean|sd|min|max)$"
  )
```

## Next EDA Tasks After BPS Is Available

- Add category CPI charts once BPS CPI table IDs are mapped.
- Add rice/fuel/electricity/LPG/rent price charts once official annual sources are ingested.
- Replace provisional distribution shares with BPS/SUSENAS expenditure-only distribution if available.
- Add wage and informal employment charts once BPS or ILOSTAT source series are confirmed.
