---
title: "Indonesia Inequality Project: EDA"
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

This Quarto notebook mirrors the Python EDA with a story-first structure. The main section keeps only the charts needed for the current narrative. Validation checks and supporting charts are kept in the appendix.

```{r setup}
library(dplyr)
library(ggplot2)
library(purrr)
library(readr)
library(scales)
library(stringr)
library(tibble)
library(tidyr)

project_root <- if (basename(getwd()) == "notebooks") dirname(getwd()) else getwd()
data_dir <- file.path(project_root, "data", "processed")

palette <- c(
  blue = "#2F6B9A",
  red = "#C44E52",
  green = "#4C956C",
  gold = "#D4A017",
  purple = "#7D5BA6",
  gray = "#6C757D",
  light = "#F6F7F9",
  dark = "#1F2933"
)

chart_theme <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", size = 14, color = palette[["dark"]]),
      plot.subtitle = element_text(size = 10, color = palette[["gray"]]),
      axis.title = element_text(size = 10, color = palette[["dark"]]),
      axis.text = element_text(size = 9, color = palette[["dark"]]),
      legend.title = element_text(size = 9, face = "bold"),
      legend.position = "bottom",
      panel.grid.minor = element_blank()
    )
}

read_processed_csv <- function(filename) {
  read_csv(file.path(data_dir, filename), show_col_types = FALSE) %>%
    rename_with(str_trim)
}
```

## Load Data

```{r load-data}
files <- c(
  gdp = "gdp_indonesia.csv",
  inequality = "inequality_indonesia.csv",
  welfare_distribution = "welfare_distribution_indonesia.csv",
  comparative_growth = "comparative_inequality_growth.csv",
  labor = "labor_indonesia.csv",
  prices = "prices_indonesia.csv"
)

data <- files %>%
  imap(~ read_processed_csv(.x)) %>%
  set_names(names(files))

indonesia_data <- data[names(data) != "comparative_growth"]
```

# Main Story

The storyline is: Indonesia is perceived as increasingly unequal, real GDP per capita has risen, average consumption improved across broad groups, but the distribution of gains shifted toward higher-income groups. The final section asks whether households feel better off once wages are compared with prices.

## Real GDP Per Capita

```{r gdp-per-capita, fig.width=10, fig.height=5.5}
gdp_per_capita <- data$gdp %>%
  drop_na(real_gdp_per_capita_lcu) %>%
  arrange(year)

gdp_per_capita_label <- gdp_per_capita %>%
  mutate(first_gdp_per_capita = first(real_gdp_per_capita_lcu)) %>%
  slice_tail(n = 1) %>%
  mutate(
    label_x = year + 0.45,
    growth_label = str_glue("+{percent(real_gdp_per_capita_lcu / first_gdp_per_capita - 1, accuracy = 1)}")
  )

gdp_per_capita %>%
  ggplot(aes(year, real_gdp_per_capita_lcu)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = gdp_per_capita_label,
    aes(x = label_x, label = growth_label),
    hjust = 0,
    size = 3.2
  ) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
  scale_x_continuous(
    breaks = seq(2000, 2024, by = 5),
    limits = c(2000, 2025.5),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  labs(
    title = "Real GDP per Capita",
    subtitle = "World Bank API, constant local currency; price effects already removed",
    x = "Year", y = "Constant Indonesian rupiah"
  ) +
  coord_cartesian(clip = "off") +
  chart_theme() +
  theme(plot.margin = margin(5.5, 55, 5.5, 5.5))
```

## Nominal Wages, Prices, And Purchasing Power

```{r wage-vs-cpi-index, fig.width=10, fig.height=5.5}
wage_price_start_year <- 2000

wage_price_index <- data$labor %>%
  select(year, avg_wage) %>%
  left_join(data$prices %>% select(year, cpi_total), by = "year") %>%
  filter(year >= wage_price_start_year) %>%
  drop_na(avg_wage, cpi_total) %>%
  mutate(real_wage = avg_wage / cpi_total * 100) %>%
  mutate(
    nominal_wage_index = avg_wage / first(avg_wage) * 100,
    cpi_index = cpi_total / first(cpi_total) * 100,
    real_wage_index = real_wage / first(real_wage) * 100
  ) %>%
  select(year, nominal_wage_index, cpi_index, real_wage_index) %>%
  pivot_longer(
    cols = ends_with("_index"),
    names_to = "metric",
    values_to = "index"
  ) %>%
  mutate(metric = recode(
    metric,
    nominal_wage_index = "Nominal wage",
    cpi_index = "Consumer prices",
    real_wage_index = "Real wage"
  ))

wage_price_index_labels <- wage_price_index %>%
  group_by(metric) %>%
  arrange(year, .by_group = TRUE) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    label_x = year + 0.28,
    label_y = case_when(
      metric == "Nominal wage" ~ index + 8,
      metric == "Consumer prices" ~ index - 2,
      metric == "Real wage" ~ index - 2,
      TRUE ~ index
    ),
    label = str_glue("{metric}  +{percent(index / 100 - 1, accuracy = 1)}")
  )

wage_price_index %>%
  ggplot(aes(year, index, color = metric)) +
  geom_hline(yintercept = 100, color = palette[["gray"]], linetype = "dashed") +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = wage_price_index_labels,
    aes(x = label_x, y = label_y, label = label, color = metric),
    hjust = 0,
    size = 3,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(
    "Nominal wage" = palette[["blue"]],
    "Consumer prices" = palette[["red"]],
    "Real wage" = palette[["green"]]
  )) +
  scale_x_continuous(
    breaks = seq(wage_price_start_year, 2024, by = 4),
    limits = c(wage_price_start_year, 2026.3),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  labs(
    title = "Nominal Wages, Prices, and Purchasing Power",
    subtitle = str_glue("Indexed to {wage_price_start_year} = 100; real wage is nominal wage divided by CPI"),
    x = "Year", y = str_glue("Index, {wage_price_start_year} = 100"), color = "Metric"
  ) +
  coord_cartesian(clip = "off") +
  chart_theme() +
  theme(
    legend.position = "none",
    plot.margin = margin(5.5, 135, 5.5, 5.5)
  )
```

## Average Consumption By Decile

```{r decile-consumption-prep, include=FALSE}
decile_levels <- c(
  "0 - 10%", "11 - 20%", "21 - 30%", "31 - 40%", "41 - 50%",
  "51 - 60%", "61 - 70%", "71 - 80%", "81 - 90%", "91 - 100%"
)

pip_percentiles <- read_csv(
  file.path(project_root, "data", "raw", "worldbank_pip_percentiles_indonesia.csv"),
  show_col_types = FALSE
) %>%
  filter(
    country_code == "IDN",
    reporting_level == "national"
  )

latest_available_consumption_year <- max(pip_percentiles$year, na.rm = TRUE)
consumption_start_year <- 2020
consumption_end_year <- min(latest_available_consumption_year, 2024)

ppp_base_year <- 2017

ppp_2017_conversion <- data$gdp %>%
  filter(year == ppp_base_year) %>%
  pull(ppp_private_consumption)

cpi_2017 <- data$prices %>%
  filter(year == ppp_base_year) %>%
  pull(cpi_total)

latest_cpi <- data$prices %>%
  filter(year <= consumption_end_year, !is.na(cpi_total)) %>%
  slice_max(year, n = 1, with_ties = FALSE)

latest_cpi_year <- latest_cpi$year
latest_price_conversion <- ppp_2017_conversion * latest_cpi$cpi_total / cpi_2017

decile_consumption <- pip_percentiles %>%
  filter(between(year, consumption_start_year, consumption_end_year)) %>%
  mutate(
    percentile = as.integer(percentile),
    decile = case_when(
      between(percentile, 1, 10) ~ "0 - 10%",
      between(percentile, 11, 20) ~ "11 - 20%",
      between(percentile, 21, 30) ~ "21 - 30%",
      between(percentile, 31, 40) ~ "31 - 40%",
      between(percentile, 41, 50) ~ "41 - 50%",
      between(percentile, 51, 60) ~ "51 - 60%",
      between(percentile, 61, 70) ~ "61 - 70%",
      between(percentile, 71, 80) ~ "71 - 80%",
      between(percentile, 81, 90) ~ "81 - 90%",
      between(percentile, 91, 100) ~ "91 - 100%"
    )
  ) %>%
  drop_na(decile, avg_welfare, pop_share) %>%
  group_by(year, decile) %>%
  summarise(
    avg_consumption = weighted.mean(avg_welfare, pop_share),
    .groups = "drop"
  ) %>%
  mutate(decile = factor(decile, levels = decile_levels)) %>%
  arrange(decile, year)

```

```{r decile-consumption-long-horizon, include=FALSE}
long_horizon_start_year <- 2000
long_horizon_end_year <- min(latest_available_consumption_year, 2024)

long_horizon_cpi <- data$prices %>%
  filter(year <= long_horizon_end_year, !is.na(cpi_total)) %>%
  slice_max(year, n = 1, with_ties = FALSE)

long_horizon_cpi_year <- long_horizon_cpi$year
long_horizon_price_conversion <- ppp_2017_conversion * long_horizon_cpi$cpi_total / cpi_2017

decile_consumption_long <- pip_percentiles %>%
  filter(between(year, long_horizon_start_year, long_horizon_end_year)) %>%
  mutate(
    percentile = as.integer(percentile),
    decile = case_when(
      between(percentile, 1, 10) ~ "0 - 10%",
      between(percentile, 11, 20) ~ "11 - 20%",
      between(percentile, 21, 30) ~ "21 - 30%",
      between(percentile, 31, 40) ~ "31 - 40%",
      between(percentile, 41, 50) ~ "41 - 50%",
      between(percentile, 51, 60) ~ "51 - 60%",
      between(percentile, 61, 70) ~ "61 - 70%",
      between(percentile, 71, 80) ~ "71 - 80%",
      between(percentile, 81, 90) ~ "81 - 90%",
      between(percentile, 91, 100) ~ "91 - 100%"
    )
  ) %>%
  drop_na(decile, avg_welfare, pop_share) %>%
  group_by(year, decile) %>%
  summarise(
    avg_consumption = weighted.mean(avg_welfare, pop_share),
    .groups = "drop"
  ) %>%
  mutate(decile = factor(decile, levels = decile_levels)) %>%
  arrange(decile, year)
```

```{r decile-consumption-index-lines, fig.width=10, fig.height=6}
cpi_index <- data$prices %>%
  filter(between(year, long_horizon_start_year, long_horizon_end_year)) %>%
  arrange(year) %>%
  mutate(cpi_index_2000 = cpi_total / first(cpi_total) * 100) %>%
  select(year, cpi_index_2000)

decile_consumption_index <- decile_consumption_long %>%
  group_by(decile) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(real_consumption_index_2000 = avg_consumption / first(avg_consumption) * 100) %>%
  ungroup() %>%
  left_join(cpi_index, by = "year") %>%
  transmute(
    year,
    series = as.character(decile),
    series_type = "Nominal decile consumption",
    index_2000 = real_consumption_index_2000 * cpi_index_2000 / 100
  )

cpi_index_line <- cpi_index %>%
  transmute(
    year,
    series = "Total CPI",
    series_type = "Total CPI",
    index_2000 = cpi_index_2000
  )

consumption_cpi_index_lines <- bind_rows(decile_consumption_index, cpi_index_line) %>%
  mutate(
    series = factor(series, levels = c(decile_levels, "Total CPI")),
    series_type = factor(series_type, levels = c("Nominal decile consumption", "Total CPI"))
  )

index_line_palette <- c(
  set_names(hcl.colors(length(decile_levels), palette = "Dark 3"), decile_levels),
  "Total CPI" = palette[["red"]]
)

consumption_cpi_index_labels <- consumption_cpi_index_lines %>%
  group_by(series) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  arrange(index_2000) %>%
  mutate(
    label_x = long_horizon_end_year + 0.28,
    label_y = case_when(
      series == "0 - 10%" ~ 930,
      series == "11 - 20%" ~ 960,
      series == "21 - 30%" ~ 990,
      series == "31 - 40%" ~ 1020,
      series == "41 - 50%" ~ 1050,
      series == "51 - 60%" ~ 1080,
      series == "61 - 70%" ~ 1110,
      series == "71 - 80%" ~ 1140,
      series == "81 - 90%" ~ 1175,
      series == "91 - 100%" ~ 1210,
      TRUE ~ index_2000
    ),
    label = str_replace_all(as.character(series), " - ", "-")
  )

consumption_cpi_index_lines %>%
  ggplot(aes(year, index_2000, color = series, linewidth = series_type, alpha = series_type, linetype = series_type)) +
  geom_hline(yintercept = 100, color = palette[["gray"]], linetype = "dashed") +
  geom_line() +
  geom_text(
    data = consumption_cpi_index_labels,
    aes(x = label_x, y = label_y, label = label, color = series),
    hjust = 0,
    size = 2.8,
    show.legend = FALSE
  ) +
  scale_x_continuous(
    breaks = seq(2000, long_horizon_end_year, by = 4),
    limits = c(long_horizon_start_year, long_horizon_end_year + 1.7),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_y_continuous(labels = label_number(accuracy = 1)) +
  scale_color_manual(values = index_line_palette) +
  scale_linewidth_manual(values = c("Nominal decile consumption" = 0.75, "Total CPI" = 1.15)) +
  scale_alpha_manual(values = c("Nominal decile consumption" = 0.82, "Total CPI" = 1)) +
  scale_linetype_manual(values = c("Nominal decile consumption" = "dashed", "Total CPI" = "solid")) +
  labs(
    title = "Nominal Average Consumption by Decile Indexed to 2000",
    subtitle = str_glue("PIP real consumption welfare converted to implied nominal indexes with total CPI; each series rebased to {long_horizon_start_year} = 100."),
    x = "Year", y = str_glue("Nominal index, {long_horizon_start_year} = 100"), color = "Series"
  ) +
  coord_cartesian(clip = "off") +
  chart_theme() +
  theme(
    legend.position = "none",
    plot.margin = margin(5.5, 78, 5.5, 5.5)
  )
```

## Distribution Shares By Population Group

```{r distribution-shares, fig.width=10, fig.height=5.5}
broad_groups <- c("Bottom 40%", "Middle 40%", "Top 20%")

ineq_long <- data$welfare_distribution %>%
  filter(group %in% broad_groups) %>%
  mutate(
    group = factor(group, levels = broad_groups),
    share = welfare_share * 100
  ) %>%
  drop_na(share)

ineq_change_labels <- ineq_long %>%
  group_by(group) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(first_share = first(share)) %>%
  slice_tail(n = 1) %>%
  ungroup() %>%
  mutate(
    change_pp = share - first_share,
    change_label = sprintf("%+.1f pp", change_pp)
  )

ineq_long %>%
  ggplot(aes(year, share, color = group)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 1.8) +
  geom_text(
    data = ineq_change_labels,
    aes(label = change_label),
    nudge_x = 0.7,
    hjust = 0,
    show.legend = FALSE
  ) +
  scale_color_manual(values = c(palette[["green"]], palette[["blue"]], palette[["red"]])) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 5), limits = c(2000, 2028)) +
  coord_cartesian(clip = "off") +
  labs(
    title = "Distribution Shares by Population Group",
    subtitle = "End labels show percentage-point change from first available year to final year",
    x = "Year", y = "Share (%)", color = "Group"
  ) +
  chart_theme() +
  theme(plot.margin = margin(5.5, 70, 5.5, 5.5))
```

```{r lower-income-welfare-price-pressure, fig.width=9, fig.height=5}
decile_consumption %>%
  group_by(decile) %>%
  arrange(year, .by_group = TRUE) %>%
  summarise(
    consumption_2020 = first(avg_consumption),
    latest_consumption = last(avg_consumption),
    growth = latest_consumption / consumption_2020 - 1,
    .groups = "drop"
  ) %>%
  mutate(
    latest_consumption_rupiah = latest_consumption * latest_price_conversion,
    consumption_label = str_glue("Rp {number(latest_consumption_rupiah, accuracy = 100, big.mark = ',')} / day"),
    label_vjust = if_else(growth >= 0, -0.35, 1.25)
  ) %>%
  ggplot(aes(decile, growth)) +
  geom_hline(yintercept = 0, color = palette[["gray"]], linetype = "dashed") +
  geom_col(fill = palette[["green"]], width = 0.72) +
  geom_text(
    aes(label = consumption_label, vjust = label_vjust),
    size = 3
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0.08, 0.14))
  ) +
  labs(
    title = "Average Consumption Growth by Decile",
    subtitle = str_glue("Bar height shows {consumption_start_year}-{consumption_end_year} growth in PIP's constant 2017 PPP consumption welfare."),
    x = "Population decile", y = "Consumption growth",
    caption = str_glue("Labels show latest consumption translated to rupiah/day in {latest_cpi_year} prices: 2017 PPP welfare x 2017 private-consumption PPP x CPI growth since 2017.")
  ) +
  chart_theme() +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1),
    legend.position = "none"
  )
```

## Household Price Pressure

```{r household-price-pressure, fig.width=9, fig.height=6}
household_analysis_start_year <- 2020
household_analysis_end_year <- 2023

category_cpi <- data$prices %>%
  select(year, cpi_food, cpi_transport, cpi_housing, cpi_education, cpi_healthcare, cpi_restaurant) %>%
  pivot_longer(
    cols = starts_with("cpi_"),
    names_to = "category",
    values_to = "index"
  ) %>%
  drop_na(index) %>%
  filter(year >= household_analysis_start_year) %>%
  mutate(category = recode(
    category,
    cpi_food = "Food, beverages, tobacco",
    cpi_transport = "Transport",
    cpi_housing = "Housing, water, electricity, fuel",
    cpi_education = "Education",
    cpi_healthcare = "Health",
    cpi_restaurant = "Restaurants"
  ))

national_wage_pressure <- data$labor %>%
  filter(between(year, household_analysis_start_year, household_analysis_end_year)) %>%
  select(year, avg_wage) %>%
  arrange(year) %>%
  drop_na(avg_wage) %>%
  mutate(
    metric = "National average wage",
    value = avg_wage
  ) %>%
  select(year, metric, value)

price_pressure <- bind_rows(
  national_wage_pressure,
  category_cpi %>%
    transmute(year, metric = category, value = index)
) %>%
  filter(between(year, household_analysis_start_year, household_analysis_end_year)) %>%
  drop_na(value) %>%
  mutate(metric = factor(
    metric,
    levels = c(
      "National average wage",
      "Food, beverages, tobacco",
      "Transport",
      "Housing, water, electricity, fuel",
      "Education",
      "Health",
      "Restaurants"
    )
  )) %>%
  group_by(metric) %>%
  arrange(year, .by_group = TRUE) %>%
  mutate(
    base_value = first(value),
    index_start = value / base_value * 100,
    emphasis = if_else(metric == "National average wage", "highlight", "muted"),
    line_type = if_else(emphasis == "highlight", "solid", "dashed")
  ) %>%
  ungroup()

price_pressure_growth <- price_pressure %>%
  group_by(metric) %>%
  arrange(year, .by_group = TRUE) %>%
  summarise(
    growth_pct = last(value) / first(value) - 1,
    .groups = "drop"
  )

price_pressure_labels <- price_pressure %>%
  group_by(metric) %>%
  slice_max(year, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(price_pressure_growth, by = "metric") %>%
  mutate(
    label_x = case_when(
      metric == "Health" ~ household_analysis_end_year + 0.76,
      metric == "Education" ~ household_analysis_end_year + 0.12,
      metric == "Housing, water, electricity, fuel" ~ household_analysis_end_year + 0.44,
      TRUE ~ household_analysis_end_year + 0.06
    ),
    label_y = index_start,
    label = str_glue("{metric}  {percent(growth_pct, accuracy = 0.1)}")
  )

price_pressure %>%
  ggplot(aes(year, index_start, group = metric)) +
  geom_hline(yintercept = 100, color = palette[["gray"]], linetype = "dashed") +
  geom_line(aes(color = metric, linetype = line_type, alpha = emphasis), linewidth = 1.05) +
  geom_point(aes(color = metric, alpha = emphasis), size = 1.8) +
  geom_text(
    data = price_pressure_labels,
    aes(x = label_x, y = label_y, label = label, color = metric),
    hjust = 0,
    size = 2.9,
    show.legend = FALSE
  ) +
  scale_x_continuous(
    breaks = household_analysis_start_year:household_analysis_end_year,
    limits = c(household_analysis_start_year, household_analysis_end_year + 1.95),
    expand = expansion(mult = c(0.01, 0.01))
  ) +
  scale_color_manual(values = c(
    "National average wage" = palette[["blue"]],
    "Food, beverages, tobacco" = palette[["green"]],
    "Transport" = palette[["gold"]],
    "Housing, water, electricity, fuel" = palette[["purple"]],
    "Education" = "#8A8F98",
    "Health" = "#A0A6AF",
    "Restaurants" = "#747B86"
  )) +
  scale_linetype_manual(values = c(solid = "solid", dashed = "dashed")) +
  scale_alpha_manual(values = c(highlight = 1, muted = 0.55)) +
  scale_y_continuous(
    breaks = seq(95, 125, by = 5),
    limits = c(95, 125)
  ) +
  labs(
    title = "How Prices Move From a Household Starting Point",
    subtitle = str_glue("National average wage and category CPI series rebased to {household_analysis_start_year} = 100 and shown through {household_analysis_end_year}."),
    x = "Year", y = "Index, first available = 100", color = "Series"
  ) +
  coord_cartesian(clip = "off") +
  chart_theme() +
  theme(
    legend.position = "none",
    plot.margin = margin(5.5, 20, 5.5, 5.5)
  )
```

## Next EDA Tasks

- Validate ILOSTAT average monthly earnings against BPS/Sakernas wage tables if a cleaner national API series is confirmed.
- Extend CPI category history before 2020 and bridge the 2024 BPS 2022=100 base only if the bridge is source-backed.
- Extend cost-of-living charts beyond wholesale rice once fuel/electricity/LPG/rent source IDs are mapped.
- Replace provisional distribution shares with BPS/SUSENAS expenditure-only distribution if available.
